---
title: "annot_plots_sr"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(ggplot2)
library(ggspavis)
library(jsonlite)
library(anndataR)
library(scuttle)
library(SpatialExperiment)
library(data.table)
library(writexl)
library(BiocParallel)
library(SingleR)
library(dplyr)
library(tidyr)
library(writexl)

# to read spatialdata objects
library(ggnewscale)
library(SpatialData)
library(SingleCellExperiment)
source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
```

# Plots post annotation of spaceranger segmentation results

Paths
```{r}
pl_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/singler_annot/spaceranger/"
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/results/SingleR_ann/spaceranger/"
```

Data
```{r}
load(paste0(Rd_path, "color_cell_dec.RData"))
load(paste0(Rd_path, "spaceranger_b1_spe.RData"))
load(paste0(Rd_path, "spe_list_nuclei_ann.RData"))
```

Select only blocco1 samples to compare with spaceranger
```{r}
spe_list_b1 <- spe_list_up[startsWith(names(spe_list_up), "blocco1")]
```


## Spatial plot of annotations

Segmentazione di spaceranger
```{r}
plots <- imap(sr_spe_list, function(spe, name) {
  plotCoords(spe, annotate = "labels_singler",
             point_size = 0.1,
             x_coord = "x", y_coord = "y") +
    scale_color_manual(values = color_vector) +
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    ) +
    ggtitle(name)
})


imap(plots, function(plot, name){
  # ggsave(paste0(pl_path,"spat_annotation/", name, "_singler_annot.png"), plot = plot, dpi = 700)
  ggsave(paste0(pl_path,"spat_annotation/", name, "_singler_annot.png"), plot = plot, dpi = 700)
})
```

la mia segmentazione
```{r}
plots <- imap(spe_list_b1, function(spe, name) {
  plotCoords(spe, annotate = "labels_singler",
             point_size = 0.1,
             x_coord = "y", y_coord = "x") +
    scale_color_manual(values = color_vector) +
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    ) +
    ggtitle(name)
})


imap(plots, function(plot, name){
  # ggsave(paste0(pl_path,"spat_annotation/", name, "_singler_annot.png"), plot = plot, dpi = 700)
  ggsave(paste0(pl_path,"spat_annotation/", name, "_b1_miei.png"), plot = plot, dpi = 700)
})
```

## Barplot

Spaceranger 
```{r}
library(forcats)
# Costruisci un data frame con tutte le proporzioni per ogni oggetto
prop_df_ordered <- imap_dfr(sr_spe_list, function(spe, name) {
  tab <- base::table(spe$labels_singler)
  prop <- prop.table(tab)
  tissue_type <- unique(spe$tissue_type)
  data.table(
    sample_id = name,
    cell_type = names(prop),
    proportion = as.numeric(prop),
    tissue_type = tissue_type
  )
}) |> 
  mutate(sample_id = fct_reorder(sample_id, tissue_type))

# orizzontale bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +
  coord_flip() +
  scale_x_discrete(name = "Sample") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +
  theme_minimal()

#ggsave(paste0(pl_path, "cell_type_proportions_oriz.png"), width = 10, height = 6, dpi = 300) # wilcoxon
ggsave(paste0(pl_path, "cell_type_proportions_spaceranger_oriz.png"), width = 10, height = 6, dpi = 300) # t-test

# vertical bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +                             # geom_col uses the y value directly (identity)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +# use your named color_vector; otherwise use scale_fill_brewer()
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

#ggsave(paste0(pl_path, "cell_type_proportions_vert.png"), width = 10, height = 6, dpi = 300)
ggsave(paste0(pl_path, "cell_type_proportions_spaceranger_vert.png"), width = 10, height = 6, dpi = 300)
```

Miei dati 
```{r}
library(forcats)
# Costruisci un data frame con tutte le proporzioni per ogni oggetto
prop_df_ordered <- imap_dfr(spe_list_b1, function(spe, name) {
  tab <- base::table(spe$labels_singler)
  prop <- prop.table(tab)
  tissue_type <- unique(spe$tissue_type)
  data.table(
    sample_id = name,
    cell_type = names(prop),
    proportion = as.numeric(prop),
    tissue_type = tissue_type
  )
}) |> 
  mutate(sample_id = fct_reorder(sample_id, tissue_type))

# orizzontale bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +
  coord_flip() +
  scale_x_discrete(name = "Sample") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +
  theme_minimal()

#ggsave(paste0(pl_path, "cell_type_proportions_oriz.png"), width = 10, height = 6, dpi = 300) # wilcoxon
ggsave(paste0(pl_path, "cell_type_proportions_b1_miei_oriz.png"), width = 10, height = 6, dpi = 300) # t-test

# vertical bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +                             # geom_col uses the y value directly (identity)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +# use your named color_vector; otherwise use scale_fill_brewer()
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

#ggsave(paste0(pl_path, "cell_type_proportions_vert.png"), width = 10, height = 6, dpi = 300)
ggsave(paste0(pl_path, "cell_type_proportions_b1_miei_vert.png"), width = 10, height = 6, dpi = 300)
```

## Boxplot

### Barplot cell type proportions stratified by sample

Spaceranger results

```{r}
prop_per_sample <- imap_dfr(sr_spe_list, function(spe, name) {
  cd <- as.data.table(SummarizedExperiment::colData(spe))
  
  cd |>
    filter(!is.na(labels_singler)) |>
    dplyr::count(labels_singler, name = "count") |>
    mutate(proportion  = count / sum(count),
           sample      = name,
           tissue_type = paste(unique(as.character(cd$tissue_type)), collapse = ", ")) |> 
    dplyr::rename(cell_type = labels_singler)
})

ggplot(prop_per_sample, aes(x = sample, y = proportion, fill = tissue_type)) +
  geom_col() + # Use geom_col() for a bar plot of existing values
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1b9e77", "c26" = "#d95f02")) +
  theme_minimal(base_size = 12) +
  theme(
    # Rotate labels 90 degrees for better readability with many samples
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(
    title = "Proportion of Cell Types per Sample",
    x = "Sample",
    y = "Proportion"
  )

ggsave(paste0(pl_path, "barplot_celltype_prop_by_sample.png"), width = 10, height = 6, dpi = 300)
#ggsave(paste0(pl_path, "boxplot_celltype_t_prop_by_cond.png"), width = 10, height = 6, dpi = 300)
```


mine segmentation results

```{r}
prop_per_sample <- imap_dfr(spe_list_b1, function(spe, name) {
  cd <- as.data.table(SummarizedExperiment::colData(spe))
  
  cd |>
    filter(!is.na(labels_singler)) |>
    dplyr::count(labels_singler, name = "count") |>
    mutate(proportion  = count / sum(count),
           sample      = name,
           tissue_type = paste(unique(as.character(cd$tissue_type)), collapse = ", ")) |> 
    dplyr::rename(cell_type = labels_singler)
})

ggplot(prop_per_sample, aes(x = sample, y = proportion, fill = tissue_type)) +
  geom_col() + # Use geom_col() for a bar plot of existing values
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1b9e77", "c26" = "#d95f02")) +
  theme_minimal(base_size = 12) +
  theme(
    # Rotate labels 90 degrees for better readability with many samples
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(
    title = "Proportion of Cell Types per Sample",
    x = "Sample",
    y = "Proportion"
  )

ggsave(paste0(pl_path, "barplot_celltype_prop_by_sample_b1_miei.png"), width = 10, height = 6, dpi = 300)
#ggsave(paste0(pl_path, "boxplot_celltype_t_prop_by_cond.png"), width = 10, height = 6, dpi = 300)
```

## Spatial relationship between trim and 2b myo

```{r}
default_color <- "white"
your_color_vector <- setNames(
  rep(default_color, 16),
  c("Endothelial", "FAPs", "Immune_Cells", "MuSC", "Myonuclei_IIx",
    "Myonuclei_IIx_IIa", "Myonuclei_IIx_IIb", "Myonuclei_MTJ",
    "Myonuclei_NMJ", "Nervous_System", "Pericyte",
    "Smooth_Muscular", "Tenocyte",
    "Myonuclei_IIb", "Myonuclei_Trim63")
)

your_color_vector["Myonuclei_IIb"]     <- "blue"
your_color_vector["Myonuclei_Trim63"]  <- "red"

plot_list <- imap(sr_spe_list, function(spe, nm) {
  plotCoords(spe, annotate = "labels_singler", point_size = 0.1) +
    scale_color_manual(values = your_color_vector, na.value = default_color) +
    ggtitle(nm) +   # <- titolo con il nome del campione
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines"),
      plot.title = element_text(hjust = 0.5)  # per centrare il titolo
    )
})
# veramente brutti così, non ha senso


plots <- imap(sr_spe_list, function(spe, nm) {
  # Determine where the annotation lives and subset accordingly
    keep <- spe$labels_singler %in% c("Myonuclei_IIb", "Myonuclei_Trim63")
    if (!any(keep)) return(NULL)
    p <- spe[, keep] |> 
      plotCoords(annotate = "labels_singler", point_size = 0.1) +
      scale_color_manual(values = your_color_vector) +
      ggtitle(nm) +   # <- titolo con il nome del campione
      theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines"),
      plot.title = element_text(hjust = 0.5)  # per centrare il titolo
      )
})

imap(plots, function(plot, name){
  #ggsave(paste0(pl_path,"spat_annotation/Trim63vsIIb/", name, ".png"), plot = plot, dpi = 700)
  ggsave(paste0(pl_path,"spat_annotation/Trim63vsIIb/", name, ".png"), plot = plot, dpi = 700)
})
```

# ----------------------------------------------------------------------------------------

# Tabelle Risultati

## Tabelle proporzioni

Per campione
```{r}
celltype_table <- imap_dfr(sr_spe_list, function(spe, name) {
  as.data.frame(colData(spe)) |> 
    filter(!is.na(labels_singler)) |>             # drop NA cell_type rows
    dplyr::count(labels_singler, name = "count")  |>      # tally per cell type
    mutate(proportion = count / sum(count),
           sample = name,
           cell_type = labels_singler)                    # use imap name as sample id
})

celltype_matrix <- celltype_table |>
  select(sample, cell_type, proportion) |>
  pivot_wider(names_from = cell_type, values_from = proportion, values_fill = 0)

celltype_matrix

# write_xlsx(celltype_matrix, paste0(res_path, "celltype_t_proportions_per_sample.xlsx"))
write_xlsx(celltype_matrix, paste0(res_path, "celltype_spaceranger_proportions_per_sample.xlsx"))
```


Per tipo di campione
```{r}
props_wide <- imap_dfr(sr_spe_list, function(spe, name) {
  data.table(
    tissue_type = rep(as.character(unique(spe$tissue_type)),
                      length.out = length(spe$labels_t_singler)),
    cell_type   = as.character(spe$labels_t_singler)
  )
}) |>
  dplyr::count(tissue_type, cell_type, name = "n") |>
  group_by(tissue_type) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  select(tissue_type, cell_type, prop) |>
  pivot_wider(names_from = cell_type, values_from = prop, values_fill = 0)

 # write_xlsx(props_wide, path = paste0(res_path, "celltype_by_tissue_type.xlsx"))
 write_xlsx(props_wide, path = paste0(res_path, "celltype_t_by_tissue_type.xlsx"))

```


# Diagnostics

## Heatmap 

```{r}
plots <- imap(ann_list, function(ann_el, name) {plotScoreHeatmap(ann_el)})

imap(plots, function(plot, name){
  #ggsave(paste0(pl_path,"spat_annotation/Trim63vsIIb/", name, ".png"), plot = plot, dpi = 700)
  ggsave(paste0(pl_path,"spat_annotation/diagnostic/heatmap/", name, ".png"), plot = plot, dpi = 700)
})

```

### Grafico heatmap dei myonuclei 
```{r}
labels_toplot <- unique(ann_list$blocco1_c26STAT3$labels)

plots <- imap(ann_list, function(ann_el, name) {
  plotScoreHeatmap(ann_el , 
                   labels.use = labels_toplot[startsWith(labels_toplot, "Myonuclei")])
  })

imap(plots, function(plot, name){
  #ggsave(paste0(pl_path,"spat_annotation/Trim63vsIIb/", name, ".png"), plot = plot, dpi = 700)
  ggsave(paste0(pl_path,"spat_annotation/diagnostic/heatmap/", name, "_myonuclei.png"), plot = plot, dpi = 700)
})
```

### Grafico heatmap senza myonuclei
```{r}
labels_toplot <- unique(ann_list$blocco1_c26STAT3$labels)

plots <- imap(ann_list, function(ann_el, name) {
  plotScoreHeatmap(ann_el , 
                   labels.use = labels_toplot[!startsWith(labels_toplot, "Myonuclei")])
  })

imap(plots, function(plot, name){
  #ggsave(paste0(pl_path,"spat_annotation/Trim63vsIIb/", name, ".png"), plot = plot, dpi = 700)
  ggsave(paste0(pl_path,"spat_annotation/diagnostic/heatmap/", name, "_altri.png"), plot = plot, dpi = 700)
})
```


## Delta distribution

```{r}

plots <- imap(ann_list, function(ann_el, name) {plotDeltaDistribution(ann_el)})

imap(plots, function(plot, name){
  #ggsave(paste0(pl_path,"spat_annotation/Trim63vsIIb/", name, ".png"), plot = plot, dpi = 700)
  ggsave(paste0(pl_path,"spat_annotation/diagnostic/delta_distribution/", name, ".png"), plot = plot, dpi = 700)
})

```

### Markers heatmap per label

```{r}
plotMarkerHeatmap(pred.ann_will, spe_list$blocco1_c26STAT3, "Myonuclei_IIx_IIb")
```

```{r}
#unique(sce_ref$subclusters)
mionuclei <- c("Myonuclei_IIb", "Myonuclei_IIx", "Myonuclei_IIx_IIb", "Myonuclei_MTJ", "Myonuclei_IIx_IIa", "Myonuclei_NMJ", "Myonuclei_Trim63")

plotMarkerHeatmap(pred.ann_will, spe_list$blocco1_c26STAT3, label = "Myonuclei_IIx_IIb",
                  other.labels = mionuclei, top = 20, num.threads = 4)
```