---
title: "confronto_emma"
output: html_document
params:
  #saves_path_stem: "my_segmentation/filtered"
  labels_key: "labels_singler"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(SpaceTrooper)
library(ggplot2)
library(ggspavis)
library(patchwork)
library(jsonlite)
library(anndataR)
library(scuttle)
library(SpatialExperiment)
library(data.table)
library(writexl)
library(BiocParallel)
library(SingleR)
library(dplyr)
library(tidyr)
library(writexl)

# to read spatialdata objects
library(ggnewscale)
#library(SpatialData)
library(SingleCellExperiment)

source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
```

# Plots post annotation of spaceranger segmentation results

Paths
```{r}
pl_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/confronto_sr_sopa/"
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/results/SingleR_ann/spaceranger/"
```

Data
```{r}
color_vector <- readRDS(paste0(Rd_path, "color_cell_dec"))
sr_list <- readRDS(paste0(Rd_path, "spaceranger_spe_nocell_expans"))
spe_list <- readRDS(paste0(Rd_path, "spe_list_nuclei_ann"))
#load(paste0(Rd_path, "spe_blocco1_emma.RData")) # dati emma coi bin
```

Select only blocco 1 and 2 samples to compare with spaceranger
```{r}
spe_list <- spe_list[names(spe_list) %in% samples_2]
names(spe_list)
spe_list

spe_list <- imap(spe_list, function(spe, name){
  keep <- which(spe$to_discard == FALSE) 
  spe[, keep]
})

sr_list <- imap(sr_list, function(spe, name){
  keep <- which(spe$to_discard == FALSE) 
  spe[, keep]
})
```

## Spatial plot of annotations

Annotazione RCTD di emma
```{r}
plots <- imap(spe_list, function(spe, name) {
  spe <- spe[, !spe$to_discard]
  spe$cell_type <- as.factor(colData(spe)[[params$labels_key]])
  plotCentroids(spe, colourBy = "cell_type", size = 1) +
  scale_color_manual(values = color_vector) +
  theme(
    legend.key.width  = unit(0.5, "lines"),
    legend.key.height = unit(1, "lines")
  ) +
  guides(color = guide_legend(override.aes = list(size = 5)))  +
  scale_y_reverse() +
  theme(
      axis.title = element_blank(), # Removes both x and y axis titles
      axis.text = element_blank(),  # Removes both x and y axis text labels
      axis.ticks = element_blank()  # Removes both x and y axis tick marks
  )
})

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"spat_annotation/", name, ".png"), 
         plot = plot, dpi = 400, units = "cm", width = 50, height = 30, create.dir = TRUE)
})
```

la mia segmentazione
```{r}
plots <- imap(spe_list, function(spe, name) {
  spe <- spe[, !spe$to_discard]
  spe$cell_type <- as.factor(colData(spe)[[params$labels_key]])
  plotCentroids(spe, colourBy = "cell_type", size = 1) +
  scale_color_manual(values = color_vector) +
  theme(
    legend.key.width  = unit(0.5, "lines"),
    legend.key.height = unit(1, "lines")
  ) +
  guides(color = guide_legend(override.aes = list(size = 5)))  +
  scale_y_reverse() +
  theme(
      axis.title = element_blank(), # Removes both x and y axis titles
      axis.text = element_blank(),  # Removes both x and y axis text labels
      axis.ticks = element_blank()  # Removes both x and y axis tick marks
  )
})

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"my_segmentation/", name, ".png"), 
         plot = plot, dpi = 400, units = "cm", width = 50, height = 30, create.dir = TRUE)
})
```

Spaceranger
```{r}
plots <- imap(sr_list, function(spe, name) {
  spe <- spe[, !spe$to_discard]
  spe$cell_type <- as.factor(colData(spe)[[params$labels_key]])
  plotCentroids(spe, colourBy = "cell_type", size = 1) +
  scale_color_manual(values = color_vector) +
  theme(
    legend.key.width  = unit(0.5, "lines"),
    legend.key.height = unit(1, "lines")
  ) +
  guides(color = guide_legend(override.aes = list(size = 5)))  +
  scale_y_reverse() +
  theme(
      axis.title = element_blank(), # Removes both x and y axis titles
      axis.text = element_blank(),  # Removes both x and y axis text labels
      axis.ticks = element_blank()  # Removes both x and y axis tick marks
  )
})

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"spaceranger/", name, ".png"), 
         plot = plot, dpi = 400, units = "cm", width = 50, height = 30, create.dir = TRUE)
})
```


## Divided by cell type
### GRID
```{r}
# solo myonuclei
plot_grids <- list()
for (spe_name in names(blocco1)) {
  spe <- blocco1[[spe_name]]
  ws <- colData(spe)[,seq(12,18,1)] # bisogna cambiare questo indicatore [,seq(8,22,1)] tutti i tipi [,seq(12,18,1)] solo myonuclei
  p <- lapply(names(ws), function(nm) {
  plotCoords(spe, annotate = nm, point_size = 0.1) +
    scale_color_gradientn(
      colors = rev(hcl.colors(9, "Rocket")),
      limits = c(0, 1)  # fissa scala da 0 a 1
    ) +
    ggtitle(nm)
  })

  final_plot <- wrap_plots(p, nrow = 3, ncol = 5, guides = "collect") &
  theme(
    legend.key.width = unit(0.5, "lines"),
    legend.key.height = unit(1, "lines"),
    strip.text = element_text(size = 5)
  ) &
  scale_color_gradientn(colors = rev(hcl.colors(9, "Rocket")), limits = c(0, 1))
  plot_grids[[spe_name]] <- final_plot
}

#grid saving
iwalk(plot_grids, function(plot, name){
  ggsave(paste0(pl_path, "spat_annotation/by_type/", name, "_myo.png"), plot = plot, dpi = 400, units = "cm", width = 50, height = 30)
})
```

my segmentation
```{r}
cell_types <- sort(names(color_vector))

# only myonuclei
(cell_types <- cell_types[startsWith(cell_types, 'Myo')])

plot_grids <- imap(spe_list, function(spe, spe_name) {
  
  # Inner loop to create a list of plots for each cell type
  list_of_plots <- map(cell_types, function(type) {
    
    # --- YOUR FIX IMPLEMENTED ---
    # 1. Create a local copy to be safe.
    spe_local <- spe
    
    # 2. Programmatically create a new column named after the content of `type`.
    # For example, if type is "Endothelial", this creates spe_local$Endothelial
    spe_local$is_here <- as.factor(colData(spe_local)[[params$labels_key]] == type)
    
    # 3. Call plotCentroids, using the `type` variable as the column name string.
    p <- SpaceTrooper::plotCentroids(spe = spe_local, colourBy = "is_here", size = 0.05) +
      scale_color_manual(values = c("TRUE" = "#CC3311", "FALSE" = "grey80")) +
      labs(title = type) + # The subplot title is the cell type name
      scale_y_reverse() +
      theme(
        axis.title = element_blank(), # Removes both x and y axis titles
        axis.text = element_blank(),  # Removes both x and y axis text labels
        axis.ticks = element_blank()  # Removes both x and y axis tick marks
      )
    
    return(p)
  })
  
  # Combine the list of plots into a single grid
  final_grid <- wrap_plots(list_of_plots, nrow = 2, ncol = 4, guides = "collect") &
    theme(
      legend.key.width = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines"),
      plot.title = element_text(size = 8)
    )
  
  # Add a main title for the entire grid
  final_grid <- final_grid + plot_annotation(title = spe_name)
  
  return(final_grid)
})

#grid saving
iwalk(plot_grids, function(plot, name){
  ggsave(paste0(pl_path, "my_segmentation/by_type_myo/", name, ".png"), plot = plot, dpi = 400, units = "cm", width = 50, height = 30, create.dir = TRUE)
})
```

Spaceranger
```{r}
cell_types <- sort(names(color_vector))

# only myonuclei
(cell_types <- cell_types[startsWith(cell_types, 'Myo')])

plot_grids <- imap(sr_list, function(spe, spe_name) {
  
  # Inner loop to create a list of plots for each cell type
  list_of_plots <- map(cell_types, function(type) {
    
    # --- YOUR FIX IMPLEMENTED ---
    # 1. Create a local copy to be safe.
    spe_local <- spe
    
    # 2. Programmatically create a new column named after the content of `type`.
    # For example, if type is "Endothelial", this creates spe_local$Endothelial
    spe_local$is_here <- as.factor(colData(spe_local)[[params$labels_key]] == type)
    
    # 3. Call plotCentroids, using the `type` variable as the column name string.
    p <- SpaceTrooper::plotCentroids(spe = spe_local, colourBy = "is_here", size = 0.05) +
      scale_color_manual(values = c("TRUE" = "#CC3311", "FALSE" = "grey80")) +
      labs(title = type) + # The subplot title is the cell type name
      scale_y_reverse() +
      theme(
        axis.title = element_blank(), # Removes both x and y axis titles
        axis.text = element_blank(),  # Removes both x and y axis text labels
        axis.ticks = element_blank()  # Removes both x and y axis tick marks
      )
    
    return(p)
  })
  
  # Combine the list of plots into a single grid
  final_grid <- wrap_plots(list_of_plots, nrow = 2, ncol = 4, guides = "collect") &
    theme(
      legend.key.width = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines"),
      plot.title = element_text(size = 8)
    )
  
  # Add a main title for the entire grid
  final_grid <- final_grid + plot_annotation(title = spe_name)
  
  return(final_grid)
})

#grid saving
iwalk(plot_grids, function(plot, name){
  ggsave(paste0(pl_path, "spaceranger/by_type_myo/", name, ".png"), plot = plot, dpi = 400, units = "cm", width = 50, height = 30, create.dir = TRUE)
})
```



## Barplot proportions

my segmentation
```{r}
library(forcats)
prop_df_ordered <- imap_dfr(spe_list, function(spe, name) {
  tab <- base::table(colData(spe)[[params$labels_key]])
  prop <- prop.table(tab)
  tissue_type <- unique(spe$tissue_type)
  data.table(
    sample_id = name,
    cell_type = names(prop),
    proportion = as.numeric(prop),
    tissue_type = as.character(tissue_type)
  )
}) |> 
  # lower case otherwise S is before f and m
  arrange(stringr::str_to_lower(tissue_type), sample_id) |>
  mutate(sample_id = fct_inorder(sample_id))

# vertical bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +                             # geom_col uses the y value directly (identity)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +# use your named color_vector; otherwise use scale_fill_brewer()
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(paste0(pl_path, "my_segmentation/cell_type_proportions_vert.png"), width = 10, height = 6, dpi = 300)
```

Spaceranger
```{r}
library(forcats)
prop_df_ordered <- imap_dfr(sr_list, function(spe, name) {
  tab <- base::table(colData(spe)[[params$labels_key]])
  prop <- prop.table(tab)
  tissue_type <- unique(spe$tissue_type)
  data.table(
    sample_id = name,
    cell_type = names(prop),
    proportion = as.numeric(prop),
    tissue_type = as.character(tissue_type)
  )
}) |> 
  # lower case otherwise S is before f and m
  arrange(stringr::str_to_lower(tissue_type), sample_id) |>
  mutate(sample_id = fct_inorder(sample_id))

# vertical bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +                             # geom_col uses the y value directly (identity)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +# use your named color_vector; otherwise use scale_fill_brewer()
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(paste0(pl_path, "spaceranger/cell_type_proportions_vert.png"), width = 10, height = 6, dpi = 300)
```



## Barplot stratified by sample

### Barplot cell type proportions stratified by sample

Spaceranger results

```{r}
prop_per_sample <- imap_dfr(blocco1, function(spe, name) {
  cd <- as.data.table(SummarizedExperiment::colData(spe))
  
  cd |>
    filter(!is.na(cell_type)) |>
    dplyr::count(cell_type, name = "count") |>
    mutate(proportion  = count / sum(count),
           sample      = name,
           tissue_type = paste(unique(as.character(cd$tissue_type)), collapse = ", "))
})

ggplot(prop_per_sample, aes(x = sample, y = proportion, fill = tissue_type)) +
  geom_col() + # Use geom_col() for a bar plot of existing values
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1b9e77", "c26" = "#d95f02")) +
  theme_minimal(base_size = 12) +
  theme(
    # Rotate labels 90 degrees for better readability with many samples
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(
    title = "Proportion of Cell Types per Sample",
    x = "Sample",
    y = "Proportion"
  )

ggsave(paste0(pl_path, "barplot_celltype_prop_by_sample_emma.png"), width = 10, height = 6, dpi = 300)
#ggsave(paste0(pl_path, "boxplot_celltype_t_prop_by_cond.png"), width = 10, height = 6, dpi = 300)
```


## Spatial relationship between trim and 2b myo

```{r}
default_color <- "white"
your_color_vector <- setNames(
  rep(default_color, 16),
  c("Endothelial", "FAPs", "Immune_Cells", "MuSC", "Myonuclei_IIx",
    "Myonuclei_IIx_IIa", "Myonuclei_IIx_IIb", "Myonuclei_MTJ",
    "Myonuclei_NMJ", "Nervous_System", "Pericyte",
    "Smooth_Muscular", "Tenocyte",
    "Myonuclei_IIb", "Myonuclei_Trim63")
)

your_color_vector["Myonuclei_IIb"]     <- "blue"
your_color_vector["Myonuclei_Trim63"]  <- "red"

plot_list <- imap(sr_spe_list, function(spe, nm) {
  plotCoords(spe, annotate = "labels_singler", point_size = 0.1) +
    scale_color_manual(values = your_color_vector, na.value = default_color) +
    ggtitle(nm) +   # <- titolo con il nome del campione
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines"),
      plot.title = element_text(hjust = 0.5)  # per centrare il titolo
    )
})
# veramente brutti così, non ha senso


plots <- imap(sr_spe_list, function(spe, nm) {
  # Determine where the annotation lives and subset accordingly
    keep <- spe$labels_singler %in% c("Myonuclei_IIb", "Myonuclei_Trim63")
    if (!any(keep)) return(NULL)
    p <- spe[, keep] |> 
      plotCoords(annotate = "labels_singler", point_size = 0.1) +
      scale_color_manual(values = your_color_vector) +
      ggtitle(nm) +   # <- titolo con il nome del campione
      theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines"),
      plot.title = element_text(hjust = 0.5)  # per centrare il titolo
      )
})

imap(plots, function(plot, name){
  #ggsave(paste0(pl_path,"spat_annotation/Trim63vsIIb/", name, ".png"), plot = plot, dpi = 700)
  ggsave(paste0(pl_path,"spat_annotation/Trim63vsIIb/", name, ".png"), plot = plot, dpi = 700)
})
```


# DOTPLOT
```{r}

df <- as.data.frame(t(expr_combat[markers, , drop=FALSE])) 
df$sample    <- colData(pb_wb7)$sample
df$condition <- colData(pb_wb7)$tissue_type
df$celltype  <- colData(pb_wb7)$cell_type

df_long <- pivot_longer(df, cols = all_of(markers),
                        names_to = "gene", values_to = "expr")

df <- df %>% filter(condition %in% c("sham","c26"))

df_long <- pivot_longer(df, cols = all_of(markers),
                        names_to = "gene", values_to = "expr")

# calcolo media e percentuale per ogni celltype × condition × gene
dot_data <- df_long %>%
  group_by(condition, celltype, gene) %>%
  summarise(
    pct_expr = sum(expr > 0) / n() * 100,
    avg_expr = mean(expr[expr > 0], na.rm = TRUE),
    .groups = "drop"
  )

# genera grafici separati per ogni marker
plots <- lapply(markers, function(g){
  ggplot(dot_data %>% filter(gene == g),
         aes(x = celltype, y = condition)) +
    geom_point(aes(size = pct_expr, color = avg_expr)) +
    scale_color_gradient(low = "grey", high = "red") +
    labs(title = g, y = "Condition", x = "Cell type",
         color = "Avg expression", size = "% expr >0") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

#salvataggio
for(i in seq_along(markers)){
  ggsave(
    filename = paste0(markers[i], "_dotplot.jpg"),  
    plot = plots[[i]],                               
    width = 8, height = 6, units = "in",            
    dpi = 300                                       
  )
}

```

#Visualizzazione nel tessuto con scale separate (no confronto)
```{r}

for (slice_name in names(subset_list)) {
  spe <- subset_list[[slice_name]]
  
  expr_matrix <- counts(spe)["Pax7", , drop = FALSE]
  
  for (gene in markers) {
    expr <- as.numeric(expr_matrix[gene, ])
    coords <- as.data.frame(spatialCoords(spe)[colnames(spe), ])
    coords$expr <- expr
    
    p <- ggplot(coords, aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = expr)) +
      geom_point(size = 0.3) +
      coord_fixed() +
      scale_y_reverse() +
      ggtitle(paste(slice_name, "-", gene)) +
      theme_void() +
      theme(plot.title = element_text(size = 8)) +
      scale_color_gradientn(colors = rev(hcl.colors(9, "Rocket")))
    
    print(p)
    
    ggsave(
      filename = file.path(sample_dir, paste0(gene, ".png")),
      plot = p,
      width = 6, height = 5, dpi = 300
    )
  }
}
```

# UMAP

```{r}
library(SpatialExperiment)
library(scater)
library(scran)

# Assuming your object is named 'spe'

# 1. Normalize counts (if not already done)
# UMAP should be run on log-transformed data, not raw counts.
spe <- logNormCounts(spe_list$blocco1_sham)

# 2. Select Highly Variable Genes (HVGs)
# Using all genes is slow and adds noise; we usually use the top ~2000 variable genes.
dec <- modelGeneVar(spe)
top_hvgs <- getTopHVGs(dec, n = 2000)

# 3. Run PCA first
# It is standard practice to run UMAP on the top Principal Components (PCs)
# rather than the full gene matrix to improve speed and denoising.
spe <- runPCA(spe, subset_row = top_hvgs)

# 4. Run UMAP
# We tell it to use the 'PCA' slot we just created.
spe <- runUMAP(spe, dimred = "PCA")

# 5. Visualization
# Color by a specific gene (e.g., "COL1A1") or a column in colData (e.g., "cluster_id")
umap_sopa <- plotUMAP(spe, colour_by = "labels_singler") 
ggsave(paste0(pl_path,"/umap_b1_sham_sopa.png"),
      plot = umap_sopa,
      width = 6, height = 5, dpi = 300)
```

spaceranger

```{r}
sr <- logNormCounts(sr_list$blocco1_sham)

# 2. Select Highly Variable Genes (HVGs)
# Using all genes is slow and adds noise; we usually use the top ~2000 variable genes.
dec <- modelGeneVar(sr)
top_hvgs <- getTopHVGs(dec, n = 2000)

# 3. Run PCA first
# It is standard practice to run UMAP on the top Principal Components (PCs)
# rather than the full gene matrix to improve speed and denoising.
sr <- runPCA(sr, subset_row = top_hvgs)

# 4. Run UMAP
# We tell it to use the 'PCA' slot we just created.
sr <- runUMAP(sr, dimred = "PCA")

# 5. Visualization
# Color by a specific gene (e.g., "COL1A1") or a column in colData (e.g., "cluster_id")
umap_sr <- plotUMAP(sr, colour_by = "labels_singler") 

ggsave(paste0(pl_path,"/umap_b1_sham_sr.png"),
      plot = umap_sr,
      width = 6, height = 5, dpi = 300)
```



