---
title: "nuclei_filtering"
output: html_document
---

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(ggplot2)
library(ggspavis)
library(ggnewscale)
library(scuttle)
library(SpatialExperiment)
library(data.table)
library(SingleR)
library(dplyr)
library(tidyr)
library(SingleCellExperiment)
library(SpaceTrooper)
library(patchwork)
source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
```

Paths
```{r}
pl_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/filtered_nuclei/"
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/results/"
```

```{r}
load(paste0(Rd_path, "spe_list_nuclei_ann.RData"))
load(paste0(Rd_path, "color_cell_dec.RData"))
# save(spe_list_up, file = paste0(Rd_path, "spe_list_nuclei_ann.RData"))
```

# Single sample analysis

```{r}
colData(spe_list_up$blocco1_c26STAT3)

spe_prova <- spe_list_up$blocco3_c26murf1
spe_prova <-  vHD_spatialPerCellQC(spe_prova)
```

Plots of QC metrics

```{r}
plotMetricHist(spe_prova, metric="sum") # total counts
plotMetricHist(spe_prova, metric="Area_um", bins = 50) # area nuclei in micron
plotMetricHist(spe_prova, metric="area", bins = 50) # are non in micron
plotMetricHist(spe_prova, metric="CountArea", bins = 50) # rapporto sum/area_um
plotMetricHist(spe_prova, metric="log2CountArea", bins = 50) # log2
plotMetricHist(spe_prova, metric="AspectRatio", bins = 50) # major axis length / minor axis length
```


Identify spatial outliers based on cell area (Area_um) and log2CountArea 

```{r}

spe_prova <- computeSpatialOutlier(spe_prova, computeBy="Area_um", method="mc")
spe_prova <- computeSpatialOutlier(spe_prova, computeBy="log2CountArea", method="mc")
#spe_prova <- computeSpatialOutlier(spe_prova, computeBy="log2AspectRatio", method="mc")
#spe_prova <- computeSpatialOutlier(spe_prova, computeBy="sum", method="mc")

names(colData(spe_prova))
base::table(Area_um = spe_prova$Area_um_outlier_mc)
base::table(CountArea = spe_prova$log2CountArea_outlier_mc)
base::table(spe_prova$log2AspectRatio_outlier_mc)
base::table(Area_outlier = spe_prova$Area_um_outlier_mc, log2countArea_outlier = spe_prova$log2CountArea_outlier_mc,
            log2AR = spe_prova$log2AspectRatio_outlier_mc)
```

To retrieve the thresholds (in the attr of the coldata column)

```{r}
attr(spe_prova$log2CountArea_outlier_mc, "thresholds")
```

```{r}
plotMetricHist(spe_prova, metric="Area_um", useFences="Area_um_outlier_mc")
#plotMetricHist(spe_prova, metric="sum", useFences="sum_outlier_mc")
plotMetricHist(spe_prova, metric="log2CountArea", useFences="log2CountArea_outlier_mc")
plotMetricHist(spe_prova, metric="CountArea")
```

Rimuovo le cellule con aree grandi (HIGH = 632) e con rapporto tra conteggio e area basso (344)
ma prima visualizziamo meglio questi nuclei prima di rimuoverli

```{r}
plotCentroids(spe_prova, colourBy = "Area_um_outlier_mc")
plotCentroids(spe_prova, colourBy = "log2CountArea_outlier_mc")
```

I nuclei con basso log2CountArea sono per lo più esterni, quindi rimuovo senza problemi, mentre i nuclei troppo grandi sembrano sparsi abbastanza casualmente nel tessuto. rimuovo anche questi. 

```{r}
table(spe_prova$log2CountArea_outlier_mc == "LOW")
table(spe_prova$Area_um_outlier_mc == "HIGH")
table(spe_prova$Area_um_outlier_mc == "HIGH" | spe_prova$log2CountArea_outlier_mc == "LOW")
spe_prova <- spe_prova[, !(spe_prova$Area_um_outlier_mc == "HIGH" | spe_prova$log2CountArea_outlier_mc == "LOW")]

```


Cell type dopo aver eliminato gli outliers

```{r}
plotCoords(spe_prova, annotate = "labels_singler",
             point_size = 0.05,
             x_coord = "x_um", y_coord = "y_um") +
    scale_color_manual(values = color_vector) +
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    )
```

```{r}
plotCoords(spe_prova, annotate = "labels_singler", point_size = 0.05)
```



--------------------------------------------------------------------------------
# All samples analysis 

Calculate QC metrics for every sample and put in the colData and the coordinate in micron instead of pixels
```{r}
spe_list_up <- imap(spe_list_up, \(spe, name) {vHD_spatialPerCellQC(spe)})
```

Plot and save most important metrics

```{r}
metric_to_plot = "Area_um"
plots <- imap(spe_list_up, \(spe, name){
  p <- plotMetricHist(spe, metric=metric_to_plot, bins = 50) +
          ggtitle(paste0(name))
  p
})

p1 <- wrap_plots(plots[c(1:12)], ncol = 4) + 
  plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt1'))
p2 <- wrap_plots(plots[c(13:24)], ncol = 4) + 
  plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt2'))

ggsave(paste0(pl_path,"metrics/", metric_to_plot, "_part1.png"), plot = p1, dpi = 300, units = "cm", width = 50, height = 30)
ggsave(paste0(pl_path,"metrics/", metric_to_plot, "_part2.png"), plot = p2, dpi = 300, units = "cm", width = 50, height = 30)
```

```{r}
# plots <- plot_metrics(spe_list_up, metric_to_plot = "sum", save = TRUE, bin = 50)
# non funziona ma non ho voglia di farla andare
```

I'll look only at the area of nuclei and the ratio between counts and area of the nuclei
Since there is only one sample that has log2CountArea asymmetric enough, I'll use sc method to determine the outliers in that metric

```{r}
spe_list_up <- imap(spe_list_up, \(spe, name){
  spe <- computeSpatialOutlier(spe, computeBy="Area_um", method="mc") # "mc" for asymmetric distribution of the choosen metric
  spe <- computeSpatialOutlier(spe, computeBy="log2CountArea", method="sc") # "sc" for symmetric distribution of the choosen metric
  spe
})
```

```{r}
base::table(spe_list_up$blocco1_c26STAT3$Area_um_outlier_mc)
base::table(spe_list_up$blocco1_c26STAT3$log2CountArea_outlier_sc)
```

```{r}
outlier_summary_df <- imap_dfr(spe_list_up, \(spe, name) {
  
  colData(spe) |>
    as.data.frame() |>
    select(Area_um_outlier_mc, log2CountArea_outlier_sc) |>
    
    # --- THE FIX IS HERE ---
    # Convert both columns to character to remove the special class
    mutate(across(everything(), as.character)) |>
    
    # Now pivot_longer will work correctly
    pivot_longer(
      cols = everything(),
      names_to = "metric",
      values_to = "outlier_type"
    ) |>
    
    count(metric, outlier_type, name = "count") |>
    
    mutate(sample = name)
})

```

```{r}
outlier_summary_df |> 
  filter(
    metric == "Area_um_outlier_mc" & outlier_type %in% c("HIGH","LOW")
  ) |> 
  select(sample, count, outlier_type, metric)


outlier_summary_df |> 
  filter(
    metric == "log2CountArea_outlier_sc" & outlier_type == "LOW"
  ) |> 
  select(sample, count, outlier_type, metric)


```

Plot and save most important metrics with the outliers thresholds

```{r}
metric_to_plot = "Area_um"
plots <- imap(spe_list_up, \(spe, name){
  p <- plotMetricHist(spe, metric=metric_to_plot, bins = 50, useFences= paste0(metric_to_plot,"_outlier_mc")) +
          labs(color="Fences outlier") +
          ggtitle(name)
  p
})

p1 <- wrap_plots(plots[c(1:12)], ncol = 4) + 
  plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt1'))
p2 <- wrap_plots(plots[c(13:24)], ncol = 4) + 
  plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt2'))

ggsave(paste0(pl_path,"outliers/", metric_to_plot, "_part1.png"), plot = p1, dpi = 300, units = "cm", width = 50, height = 30)
ggsave(paste0(pl_path,"outliers/", metric_to_plot, "_part2.png"), plot = p2, dpi = 300, units = "cm", width = 50, height = 30)
```

Plots of the nuclei centroids colored by outliers info

```{r}

# "Area_um_outlier_mc"
# "log2CountArea_outlier_sc"


metric_to_plot = "log2CountArea_outlier_sc"
plots <- imap(spe_list_up, \(spe, name){
  p <- plotCentroids(spe, colourBy = metric_to_plot, size = 0.5) +
          ggtitle(name)
  p
})

# p1 <- wrap_plots(plots[c(1:12)], ncol = 4) + 
#   plot_annotation(title = paste0('Nuclei outliers - ', metric_to_plot))
# p2 <- wrap_plots(plots[c(13:24)], ncol = 4) + 
#   plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt2'))
# 
# ggsave(paste0(pl_path,"area_outlier/", metric_to_plot, "_part1.png"), plot = p1, dpi = 300, units = "cm", width = 50, height = 30)
# ggsave(paste0(pl_path,"area_outlier/", metric_to_plot, "_part2.png"), plot = p2, dpi = 300, units = "cm", width = 50, height = 30)

iwalk(plots, function(plot, name){
 ggsave(paste0(pl_path,"CountArea_outlier/", name, ".png"), plot = plot, dpi = 300, units = "cm", width = 50, height = 30)
  #ggsave(paste0(pl_path,"area_outlier/", name, ".png"), plot = plot, dpi = 300, units = "cm", width = 50, height = 30)
})
```


# Changes in the annotation

Let's see what change in the annotation proportion if I remove the supposed outliers

```{r}
table(spe_list_up$blocco1_c26STAT3$log2CountArea_outlier_sc)
table(spe_list_up$blocco1_c26STAT3$Area_um_outlier_mc)
```

Let's fix a median value for the Area_um HIGH and LOW so to cut on every sample
at the same dimensions
```{r}
threshes <- sapply(spe_list_up, \(spe){
       attr(spe$Area_um_outlier_mc,"thresholds")
})
(LOW_MEDIAN <- median(threshes[1,]))
(HIGH_MEDIAN <- median(threshes[2,]))
```


Let's remove the log2CountArea_outlier_sc if they are LOW
```{r}
spe_list_filtered <- imap(spe_list_up, function(spe, name){
  area_out <- ((spe$Area_um < LOW_MEDIAN) | (spe$Area_um > HIGH_MEDIAN))
  ca_out <- spe$log2CountArea_outlier_sc == "LOW"
  to_remove <- area_out | ca_out
  # check
  stopifnot(length(to_remove) == ncol(spe))
  spe_f <- spe[, !to_remove]
  spe_f
})
```


```{r}
spe_list_wb7 <- spe_list_filtered[!startsWith(names(spe_list_filtered), "blocco7")]
```


## Barplot of the proportion of cells for the different annotation methods

```{r}
library(forcats)
# Costruisci un data frame con tutte le proporzioni per ogni oggetto
prop_df_ordered <- imap_dfr(spe_list_filtered, function(spe, name) {
  tab <- base::table(spe$labels_singler)
  prop <- prop.table(tab)
  tissue_type <- unique(spe$tissue_type)
  data.table(
    sample_id = name,
    cell_type = names(prop),
    proportion = as.numeric(prop),
    tissue_type = tissue_type
  )
}) |> 
  mutate(sample_id = fct_reorder(sample_id, tissue_type))

# orizzontale bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +
  coord_flip() +
  scale_x_discrete(name = "Sample") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +
  theme_minimal()

ggsave(paste0(pl_path, "cell_type_proportions_oriz.png"), width = 10, height = 6, dpi = 300) # t-test

# vertical bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +                             # geom_col uses the y value directly (identity)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +# use your named color_vector; otherwise use scale_fill_brewer()
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(paste0(pl_path, "cell_type_proportions_vert.png"), width = 10, height = 6, dpi = 300)
```

#### Barplot aggregato per condizione (SENZA BLOCCO 7)

Sia Wilcox che t-test

```{r}
prop_df_condition <- imap_dfr(spe_list_wb7, function(spe, name) {
  data.table(
    tissue_type = spe$tissue_type,
    cell_type = spe$labels_rctd
  ) |> 
    filter(!is.na(cell_type))
}) |> 
  dplyr::count(tissue_type, cell_type) |>
  group_by(tissue_type) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()


ggplot(prop_df_condition, aes(x = tissue_type, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per condizione",
    x = "Condizione (tissue_type)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(paste0(pl_path, "cell_type_rctd_proportions_by_condition.png"), width = 10, height = 6, dpi = 300)
```

