---
title: "annot_marker"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(ggplot2)
library(ggspavis)
library(jsonlite)
library(anndataR)
library(scuttle)
library(SpatialExperiment)
library(data.table)
library(SpaceTrooper)
library(SingleR)
# to read spatialdata objects
library(ggnewscale)
#library(SpatialData)
library(SingleCellExperiment)
library(SingleR)
library(BiocParallel)
source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
```

Paths
```{r}
pl_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/singler_annot/markers/"
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/results/SingleR_ann/markers/"
```

Loading spatial experiment and single-cell reference and others
```{r}
load(paste0(Rd_path, "color_cell_dec.RData"))
load(paste0(Rd_path, "spaceranger_b1_spe.RData"))
load(paste0(Rd_path, "spe_list_nuclei_ann.RData"))
(sce_ref <- readRDS(ref_path))
```

```{r}
spe_list_b1 <- spe_list_up[startsWith(names(spe_list_up), "blocco1")]
rm(spe_list_up)
spe_prova <- spe_list_b1$blocco1_c26STAT3
```

# Marker defined annotation

## List of marker genes

```{r}
unique(sce_ref$subclusters)
marker_list <- list(
  
  # Smooth Muscular Cells
  Smooth_Muscular = c("Myh11", "Trpc3", "Cacna1c"),
  
  # Myonuclei Subtypes
  Myonuclei_IIb = c("Ttn", "Myh4"),
  Myonuclei_IIx = c("Ttn", "Myh1"),
  Myonuclei_IIx_IIa = c("Ttn", "Myh1", "Myh2"), 
  Myonuclei_IIx_IIb = c("Ttn", "Myh1", "Myh4"),
  Myonuclei_MTJ = c("Ttn", "Col22a1"),
  Myonuclei_NMJ = c("Ttn", "Ache", "Etv5", "Ano4", "Chrne"),
  Myonuclei_Trim63 = c("Trim63"),
  
  # Fibro-Adipogenic Progenitors
  FAPs = c("Pdgfra", "Ly6a", "Dcn", "Ankrd1", "Slc24a2", "Adamts20"),
  
  # Endothelial Cells
  Endothelial = c("Pecam1", "Kdr"),
  
  # Muscle Stem Cells (Satellite Cells)
  MuSC = c("Pax7", "Col6a3"),
  
  # Tenocytes
  Tenocyte = c("Col11a1", "Mkx", "Scx", "Col1a1", "Itgb1", "Col24a1", "Col1a2", "Col6a1"),
  
  # Immune Cells (Combined)
  Immune_Cells = unique(c(
    # General
    "Ptprc", 
    # Macrophages
    "Cd68", "Cd74", "Itgam", "Cd14", "F13a1", "Mrc1", "Mctp1", "Slc9a9", "Rbpj",
    # T-cells
    "Cd3d", "Cd3e", "Cd3g", "Cd4",
    # B-cells
    "Cd19", "Cd79a", "Cd22", "Ms4a1"
  )),
  
  # Nervous System (Combined)
  Nervous_System = unique(c(
    # General / Glial
    "NCAM", "Ptn", "Postn", "Cadm1",
    # Schwann cells
    "Mbp", "Mpz", "Prx",
    # Neurons
    "Sfrp5", "Kcna1", "Matn2"
  )),
  
  # Pericytes
  Pericyte = c("Reln", "Mmrn1", "Mcam", "Pdgfrb")
)

# You can print the updated list to verify
# For example, let's look at the newly expanded Immune_Cells list
marker_list
```

```{r}
pred.mark <- SingleR(test=spe_prova, ref=sce_ref, labels=sce_ref$subclusters, genes=marker_list, BPPARAM=MulticoreParam(8))
```

## grafici 

Vediamo com'è andata
```{r}

plotCentroids(spe_prova, colourBy = "label_mark", size = 0.2) +
scale_color_manual(values = color_vector) +
theme(
  legend.key.width  = unit(0.5, "lines"),
  legend.key.height = unit(1, "lines")
) +
guides(color = guide_legend(override.aes = list(size = 5)))  +
scale_y_reverse()


iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"spat_annotation/", name, "_b1_miei.png"), 
         plot = plot, dpi = 400, units = "cm", width = 50, height = 30)
})
```


--------------------------------------------------------------------------------
# SingleR with multiple Reference
--------------------------------------------------------------------------------
## Reference specific labels

Columns useful in the sce_ref:
  - cell_type : subclusters
  - labels of the references : condition 

I need to build a column of subsubclusters aka labels differentiated by condition

```{r}

```


```{r}
pred.comb <- SingleR(spe_prova, ref=sce_ref, labels=labels_cond, assay.type.test=1)
table(com.res1$labels)
```

Problem with batch effects and technical artifacts



--------------------------------------------------------------------------------
## Combining inference from individual references

```{r}
table(spe_prova$condition)

sham_ref <- sce_ref[, sce_ref$condition == "SHAM"]
cac_ref <- sce_ref[, sce_ref$condition == "CAC"]


pred.comb2 <- SingleR(test = spe_prova, assay.type.test=1,
    ref = list(SPD=spd, HPCA=hpca), 
    labels = list(bpe$label.main, hpca$label.main))
```







