---
title: "Other annotation methods from SingleR"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(ggplot2)
library(ggspavis)
library(jsonlite)
library(anndataR)
library(scuttle)
library(SpatialExperiment)
library(data.table)
library(SpaceTrooper)
library(SingleR)
# to read spatialdata objects
library(ggnewscale)
#library(SpatialData)
library(SingleCellExperiment)
library(SingleR)
library(BiocParallel)
source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
```

Paths
```{r}
pl_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/singler_annot/altro/"
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/results/SingleR_ann/altro/"
```

Loading spatial experiment and single-cell reference and others
```{r}
# spaceranger - blocco1 e blocco2 no cell expansion
sr_spe_list <- readRDS(paste0(Rd_path, "spaceranger_spe_nocell_expans"))
# spaceranger - blocco 1 e blocco2 cell expansion 2 microns
#sr_spe_list <- readRDS(paste0(Rd_path, "spaceranger_spe_cell_expans_2um"))
spe_list <- readRDS(paste0(Rd_path, "spe_list_nuclei_ann"))
sce_ref <- readRDS(ref_path)
color_vector <- readRDS(paste0(Rd_path, "color_cell_dec"))
```

```{r}
spe_list_b1 <- spe_list[startsWith(names(spe_list), "blocco1")]
rm(spe_list)
spe_prova <- spe_list_b1$blocco1_sham
```

# Marker defined annotation

## List of marker genes

```{r}
unique(sce_ref$subclusters)
marker_genes <- list(
  "Smooth_Muscular" = c("Myh11", "Trpc3", "Cacna1c"),
  "Myonuclei_IIb" = c("Ttn", "Myh4"),
  "Myonuclei_IIx" = c("Ttn", "Myh1"),
  "Myonuclei_IIx_IIa" = c("Ttn", "Myh1", "Myh2"),
  "Myonuclei_IIx_IIb" = c("Ttn", "Myh1", "Myh4"),
  "Myonuclei_MTJ" = c("Ttn", "Col22a1", "Itgb1", "Col24a1", "Col1a2", "Col6a1", "Ncam"),
  "Myonuclei_NMJ" = c("Ttn", "Ache", "Etv5", "Ano4", "Chrne"),
  "Myonuclei_Trim63" = c("Trim63"),
  "FAPs" = c("Pdgfra", "Cd140a", "Sca1", "Ly6a", "Dcn", "Ankrd1", "Slc24a2", "Adamts20"),
  "Endothelial" = c("Pecam1", "Kdr"), # Changed from "Endothelial Cells"
  "MuSC" = c("Pax7", "Col6a3"),
  "Tenocyte" = c("Col11a1", "Mkx", "Scx", "Col1a1"), # Changed from "Tenocytes"
  "Immune_Cells" = c(
    "Ptprc", "Cd68", "Cd74", "Itgam", "Cd14", "F13a1", "Mrc1", 
    "Mctp1", "Slc9a9", "Rbpj", "Cd3d", "Cd3e", "Cd3g", "Cd4", 
    "Cd19", "Cd79a", "Cd22", "Ms4a1"
  ),
  "Nervous_System" = c(
    "Ptn", "Postn", "Cadm1", "Mbp", "Mpz", "Prx", "Sfrp5", 
    "Kcna1", "Matn2"
  ),
  "Pericyte" = c("Reln", "Mmrn1", "Mcam", "Pdgfrb") # Changed from "Pericytes"
)



#saveRDS(marker_genes, file = paste0(Rd_path, "marker_genes_list"))

marker_genes
```

```{r}
pred.mark <- SingleR(test=spe_prova, ref=sce_ref, labels=sce_ref$subclusters, genes=marker_genes, BPPARAM=MulticoreParam(8))
table(pred.mark$labels)
table(spe_prova$labels_singler)
```

## grafici 

Vediamo com'è andata
```{r}
spe_prova$label_marker <- as.factor(pred.mark$labels)
plotCentroids(spe_prova, colourBy = "label_marker", size = 0.2) +
scale_color_manual(values = color_vector) +
theme(
  legend.key.width  = unit(0.5, "lines"),
  legend.key.height = unit(1, "lines")
) +
guides(color = guide_legend(override.aes = list(size = 5)))  +
scale_y_reverse()


iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"spat_annotation/", name, "_b1_miei.png"), 
         plot = plot, dpi = 400, units = "cm", width = 50, height = 30)
})
```

9/11
Direi non bene... ma c'è sicuramente un errore nei markers, GPT ha sbagliato qualcosa nel rimettere assieme il file excel, attendiamo correzione.
13/11
Meglio? dalla table non sembra malaccio, c'e stata una redistribuzione ma non cosi eclatanti le differente

--------------------------------------------------------------------------------
# SingleR with multiple References
--------------------------------------------------------------------------------

There are different strategies implemented in the library to account for multiple References.

## Reference specific labels

In this strategy, each label is defined in the context of its reference dataset. This means that a label - say, “B cell” - in reference dataset X is considered to be different from a “B cell” label in reference dataset Y. Use of reference-specific labels is most appropriate if there are relevant biological differences between the references; for example, if one reference is concerned with healthy tissue while the other reference considers diseased tissue, it can be helpful to distinguish between the same cell type in different biological contexts.

We can easily implement this approach by combining the expression matrices together and pasting the reference name onto the corresponding character vector of labels. This modification ensures that the downstream SingleR() call will treat each label-reference combination as a distinct entity.

Columns useful in the sce_ref:
  - cell_type : subclusters
  - labels of the references : condition 

I need to build a column of cell type's labels unique for every condition

```{r}
head(sce_ref$subclusters)
head(sce_ref$condition)
head(paste0(sce_ref$condition, "_", sce_ref$subclusters))
sce_ref$label.main <- paste0(sce_ref$condition, "_", sce_ref$subclusters)
```

pred.multi: de.method = "classic", ovvero come fossero dati bulk
pred.multi2: de.method = "wilcox", rank test di wilcoxon per sc rna-seq reference

```{r}
# pred.multi <- SingleR(spe_prova, ref=sce_ref, labels=sce_ref$label.main, de.method = "classic", BPPARAM=MulticoreParam(8))
pred.multi2 <- SingleR(spe_prova, ref=sce_ref, labels=sce_ref$label.main, de.method = "wilcox", BPPARAM=MulticoreParam(8))
table(pred.multi$labels)
table(pred.multi2$labels)

spe_prova$label_multi <- pred.multi2$labels
```

```{r}
table(pred.multi2$labels)[grepl("^CAC_", names(table(pred.multi2$labels)))]
table(pred.multi$labels)[grepl("^CAC_", names(table(pred.multi$labels)))]
table(pred.multi2$labels)[grepl("^SHAM_", names(table(pred.multi2$labels)))]
table(pred.multi$labels)[grepl("^SHAM_", names(table(pred.multi$labels)))]
```


This method of combining different reference datasets has problems with batch effects and technical artifacts



--------------------------------------------------------------------------------
## Combining inference from individual references

Another strategy - and the default approach implemented in SingleR() - involves performing classification separately within each reference, and then collating the results to choose the label with the highest score across references. This is a relatively expedient approach that avoids the need for explicit harmonization while also reducing exposure to reference-specific batch effects.

To use this method, we simply pass multiple objects to the ref= and label= argument in SingleR(). The combining strategy is as follows:

    The function first annotates the test dataset with each reference individually in the same manner as described in Section 1.2. This step is almost equivalent to simply looping over all individual references and running SingleR() on each.
    For each cell, the function collects its predicted labels across all references. In doing so, it also identifies the union of markers that are upregulated in the predicted label in each reference.
    The function identifies the overall best-scoring label as the final prediction for that cell. This step involves a recomputation of the scores across the identified marker subset to ensure that these scores are derived from the same set of genes (and are thus comparable across references).

The function will then return a DataFrame of combined results for each cell in the test dataset, including the overall label and the reference from which it was assigned.

```{r}
comb_ref_singler <- function(spe, sce_ref, ...){
  sham_ref <- sce_ref[, sce_ref$condition == "SHAM"]
  cac_ref <- sce_ref[, sce_ref$condition == "CAC"]
  
  pred.ann <- SingleR(test = spe,
    ref = list(SHAM=sham_ref, CAC=cac_ref), 
    labels = list(sham_ref$subclusters, cac_ref$subclusters), ...)
  pred.ann
}
```

```{r}
pred.comb <- comb_ref_singler(spe_prova, sce_ref, BPPARAM=MulticoreParam(8), de.method = "wilcox")
table(pred.comb$labels) 
table(pred.comb$reference) 
table(labels = pred.comb$labels, ref = pred.comb$reference) 

spe_prova$label_comb <- pred.comb$labels
```

### Diagnostic

```{r}
plotScoreHeatmap(pred.comb)
plotDeltaDistribution(pred.comb)
```

# grafici

```{r}
plot_columns <- list("singler" = "labels_singler", 
                     "combined" = "label_comb", 
                     "multi" = "label_multi")

plots <- imap(plot_columns, function(col, name) {
  
  # spe_prova$label_multi <- spe_prova$label_multi[startsWith(spe_prova$label_multi ,"SHAM")]
  colData(spe_prova)[[col]] <- as.factor(colData(spe_prova)[[col]])
  
  # 1. Create the base plot for the current column
  p <- plotCentroids(spe_prova, colourBy = col, size = 0.5) +
  
  # 2. Add your custom layers
    scale_color_manual(values = color_vector) +
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    ) +
    guides(color = guide_legend(override.aes = list(size = 5))) +
    scale_y_reverse() +
    
  # 3. (Optional but recommended) Add a title to know which plot is which
    labs(title = paste("Centroids of ", name, " colored by: ", col))
  
  # 4. Return the complete plot object
  p
})
plots

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"spat_annotation/", name, ".png"), 
         plot = plot, dpi = 400, units = "cm", width = 50, height = 30)
})
```

```{r}
prova <- \(col = "label_multi", name = "multi") {
  
  names(color_vector) <- paste0("SHAM_",names(color_vector))
  
  spe_tmp <- spe_prova[, startsWith(spe_prova$label_multi,"SHAM")]
  spe_tmp$label_multi <- as.factor(spe_tmp$label_multi)
  
  
  # 1. Create the base plot for the current column
  p <- plotCentroids(spe_prova, colourBy = col, size = 0.5) +
  
  # 2. Add your custom layers
    scale_color_manual(values = color_vector) +
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    ) +
    guides(color = guide_legend(override.aes = list(size = 5))) +
    scale_y_reverse() +
    
  # 3. (Optional but recommended) Add a title to know which plot is which
    labs(title = paste("Centroids of ", name, " colored by: ", col))
  
  # 4. Return the complete plot object
  p
}
prova()
```



