---
title: "Plots post annotation"
output: html_document
params:
  saves_path_stem: "my_segmentation/filtered"
  labels_key: "labels_singler"
  spe_key: "filtered_spe_annotated_28Nov"
  partial_data: TRUE
  single_plots: FALSE
  subtype_plots: FALSE
  diagnostics: FALSE
  ann_key: "results_mia_segment_singleR_annot"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(ggplot2)
library(ggspavis)
library(patchwork)
library(jsonlite)
library(anndataR)
library(scuttle)
library(SpatialExperiment)
library(data.table)
library(writexl)
library(BiocParallel)
library(SingleR)
library(dplyr)
library(tidyr)
library(writexl)
library(SingleCellExperiment)
library(ggnewscale)
library(SpaceTrooper)
library(forcats)

source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
```

# Plots post annotation

Paths
```{r}
pl_path = paste0("/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/SingleR_annot/", params$saves_path_stem,"/")
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = paste0("/mnt/europa/valerio/repositories/cachetic_visiumHD/results/SingleR_annot/", params$saves_path_stem, "/")
```

Data
```{r}
color_vector <- readRDS(paste0(Rd_path, "color_cell_dec"))
spe_list <- readRDS(paste0(Rd_path, params$spe_key))
#spe_list <- readRDS(paste0(Rd_path, "spe_list_nuclei_ann"))
# spe_list <- spe_list[stringr::str_detect(names(spe_list), "^blocco[12]")]
# saveRDS(filtered_spe, file = paste0(Rd_path, "filtered_spe_annotated_28Nov"))
# spe_list
```

Filtering bad tissue areas

```{r}
spe_list <- spe_list |> 
  map(function(spe) {
    spe <- spe[, !spe$to_discard]
    spe
  })
```


## Spatial visualization of the annotated spots/nuclei

```{r}
plots <- imap(spe_list, function(spe, name) {
  spe <- spe[, !spe$to_discard]
  spe$cell_type <- as.factor(colData(spe)[[params$labels_key]])
  plotCentroids(spe, colourBy = "cell_type", size = 1) +
  scale_color_manual(values = color_vector) +
  theme(
    legend.key.width  = unit(0.5, "lines"),
    legend.key.height = unit(1, "lines")
  ) +
  guides(color = guide_legend(override.aes = list(size = 5)))  +
  scale_y_reverse() +
  theme(
      axis.title = element_blank(), # Removes both x and y axis titles
      axis.text = element_blank(),  # Removes both x and y axis text labels
      axis.ticks = element_blank()  # Removes both x and y axis tick marks
  )
})

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"spat_annotation/", name, ".png"), 
         plot = plot, dpi = 400, units = "cm", width = 50, height = 30, create.dir = TRUE)
})
#singles
# ggsave(filename = paste0(pl_path,"plot_emma_b2_c26.png"), 
#          plot = plots$blocco2_c26, dpi = 400, units = "cm", width = 50, height = 30)
```


## Divided by cell type

Myo
```{r}
cell_types <- sort(names(color_vector))

# only myonuclei
(cell_types <- cell_types[startsWith(cell_types, 'Myo')])

plot_grids <- imap(spe_list, function(spe, spe_name) {
  
  # Inner loop to create a list of plots for each cell type
  list_of_plots <- map(cell_types, function(type) {
    
    # --- YOUR FIX IMPLEMENTED ---
    # 1. Create a local copy to be safe.
    spe_local <- spe
    
    # 2. Programmatically create a new column named after the content of `type`.
    # For example, if type is "Endothelial", this creates spe_local$Endothelial
    spe_local$is_here <- as.factor(colData(spe_local)[[params$labels_key]] == type)
    
    # 3. Call plotCentroids, using the `type` variable as the column name string.
    p <- SpaceTrooper::plotCentroids(spe = spe_local, colourBy = "is_here") +
      scale_color_manual(values = c("TRUE" = "#CC3311", "FALSE" = "grey80")) +
      labs(title = type) + # The subplot title is the cell type name
      scale_y_reverse() +
      theme(
        axis.title = element_blank(), # Removes both x and y axis titles
        axis.text = element_blank(),  # Removes both x and y axis text labels
        axis.ticks = element_blank()  # Removes both x and y axis tick marks
      )
    
    return(p)
  })
  
  # Combine the list of plots into a single grid
  final_grid <- wrap_plots(list_of_plots, nrow = 2, ncol = 4, guides = "collect") &
    theme(
      legend.key.width = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines"),
      plot.title = element_text(size = 8)
    )
  
  # Add a main title for the entire grid
  final_grid <- final_grid + plot_annotation(title = spe_name)
  
  return(final_grid)
})

#grid saving
iwalk(plot_grids, function(plot, name){
  ggsave(paste0(pl_path, "spat_annotation/by_type/", name, ".png"), plot = plot, dpi = 400, units = "cm", width = 50, height = 30, create.dir = TRUE)
})
```


### SINGLE plots
```{r conditional-plot-chunk, eval = params$single_plots}
# This entire chunk will only be evaluated if params$single_plots is TRUE.

cell_types <- sort(names(color_vector))

# only myonuclei
# (cell_types <- cell_types[startsWith(cell_types, 'Myonuclei')])

plot_lists <- imap(spe_list, function(spe, name) {
  
  # for each sample create a plot of every cell type
  
  # Inner loop to create a list of plots for each cell type
  list_of_plots <- map(cell_types, function(type) {
    
    # --- YOUR FIX IMPLEMENTED ---
    # 1. Create a local copy to be safe.
    spe_local <- spe
    
    # 2. Programmatically create a new column named after the content of `type`.
    # For example, if type is "Endothelial", this creates spe_local$Endothelial
    spe_local$is_here <- as.factor(colData(spe_local)[[params$labels_key]] == type)
    
    # 3. Call plotCoords, using the `type` variable as the column name string.
    p <- SpaceTrooper::plotCentroids(spe = spe_local, colourBy = "is_here", size = 1.5) + 
      scale_color_manual(values = c("TRUE" = "#CC3311", "FALSE" = "grey")) +
      labs(title = type) + # The subplot title is the cell type name 
      scale_y_reverse() +
      theme(
        axis.title = element_blank(), # Removes both x and y axis titles
        axis.text = element_blank(),  # Removes both x and y axis text labels
        axis.ticks = element_blank()  # Removes both x and y axis tick marks
      )

    
    return(p)
  })
  names(list_of_plots) <- cell_types

  return(list_of_plots)
})


iwalk(plot_lists, function(sample, name){
  iwalk(sample, function(celltype, cellname){
    ggsave(paste0(pl_path, "myonuclei/", name,"/", cellname, ".png"), plot = celltype, dpi = 400, units = "cm", width = 50, height = 30, create.dir = TRUE)
  })
})

#singles
# ggsave(filename = paste0(pl_path,"plot_emma_b2_c26_myotrim63.png"), 
#          plot = plot_save , dpi = 400, units = "cm", width = 50, height = 30)
```

## Spatial relationship between specific cell types

```{r second-conditional-plot-chunk, eval = params$subtype_plots}

default_color <- "grey80"
your_color_vector <- setNames(
  rep(default_color, 16),
  c("Endothelial", "FAPs", "Immune_Cells", "MuSC", "Myonuclei_IIx",
    "Myonuclei_IIx_IIa", "Myonuclei_IIx_IIb", "Myonuclei_MTJ",
    "Myonuclei_NMJ", "Nervous_System", "Pericyte",
    "Smooth_Muscular", "Tenocyte",
    "Myonuclei_IIb", "Myonuclei_Trim63")
)

cell_types_to_keep <- c("Myonuclei_IIb", "Myonuclei_Trim63")

your_color_vector[cell_types_to_keep[1]] <- "red3"
your_color_vector[cell_types_to_keep[2]] <- "blue"


plots <- imap(spe_list, function(spe, name){
  spe$cell_type <- colData(spe)[[params$labels_key]]
  # If a value in 'cell_type' is in our 'keep' list, we keep it. Otherwise, it becomes NA.
  spe$cell_type <- as.factor(ifelse(spe$cell_type %in% cell_types_to_keep, # condition
    as.character(spe$cell_type), # if is true: Keep the original value (as character)
    NA                                # otherwise: Replace with NA
  ))
  
  plotCentroids(spe, colourBy = "cell_type", size = 0.3) +
    scale_color_manual(values = your_color_vector, na.value = default_color) +
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    ) +
    guides(color = guide_legend(override.aes = list(size = 5)))  +
    scale_y_reverse() +
    theme(
      axis.title = element_blank(), # Removes both x and y axis titles
      axis.text = element_blank(),  # Removes both x and y axis text labels
      axis.ticks = element_blank()  # Removes both x and y axis tick marks
    )
})

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"spat_annotation/Trim63vsIIb/", name, ".png"), plot = plot, dpi = 400, units = "cm", width = 50, height = 30, create.dir = TRUE)
})
```

# Barplots

## Proporzione dei tipi cellulari divisi per campione

```{r}
# Costruisci un data frame con tutte le proporzioni per ogni oggetto
prop_df_ordered <- imap_dfr(spe_list, function(spe, name) {
  tab <- base::table(colData(spe)[[params$labels_key]])
  prop <- prop.table(tab)
  tissue_type <- unique(spe$tissue_type)
  data.table(
    sample_id = name,
    cell_type = names(prop),
    proportion = as.numeric(prop),
    tissue_type = as.character(tissue_type)
  )
}) |> 
  # lower case otherwise S is before f and m
  arrange(stringr::str_to_lower(tissue_type), sample_id) |>
  mutate(sample_id = fct_inorder(sample_id))

# vertical bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +                             # geom_col uses the y value directly (identity)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +# use your named color_vector; otherwise use scale_fill_brewer()
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(paste0(pl_path, "cell_type_proportions_vert.png"), width = 10, height = 6, dpi = 300)
```

#### Proporzioni aggregate per tipo di tessuto

```{r}
prop_df_condition <- imap_dfr(spe_list, function(spe, name) {
  data.table(
    tissue_type = as.character(spe$tissue_type),
    cell_type = colData(spe)[[params$labels_key]]
  ) |> 
    filter(!is.na(cell_type))
}) |> 
  arrange(stringr::str_to_lower(tissue_type)) |> 
  dplyr::count(tissue_type, cell_type) |>
  group_by(tissue_type) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()


ggplot(prop_df_condition, aes(x = tissue_type, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per condizione",
    x = "Condizione (tissue_type)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(paste0(pl_path, "cell_type_proportions_by_condition.png"), width = 10, height = 6, dpi = 300)
```

### Barplot cell type proportions stratified by sample

If not full dataset I can't do boxplots

```{r barplot-insteadof-boxplot, eval = params$partial_data}
# This will run if partial_data == TRUE

prop_per_sample <- imap_dfr(spe_list, function(spe, name) {
  cd <- as.data.table(SummarizedExperiment::colData(spe))
  cd |>
    filter(!is.na(.data[[params$labels_key]])) |>
    dplyr::count(.data[[params$labels_key]], name = "count") |>
    mutate(proportion  = count / sum(count),
           sample      = name,
           tissue_type = paste(unique(as.character(cd$tissue_type)), collapse = ", ")) |> 
    dplyr::rename(cell_type = all_of(params$labels_key))
}) |> 
  # 1. First, arrange the entire dataframe by tissue_type, then by sample_id
  arrange(tissue_type, sample) |>
  # 2. Then, create a factor from sample_id that follows its current order
  mutate(sample = fct_inorder(sample))
  

ggplot(prop_per_sample, aes(x = sample, y = proportion, fill = tissue_type)) +
  geom_col() + # Use geom_col() for a bar plot of existing values
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1b9e77", "c26" = "#d95f02")) +
  theme_minimal(base_size = 12) +
  theme(
    # Rotate labels 90 degrees for better readability with many samples
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(
    title = "Proportion of Cell Types per Sample",
    x = "Sample",
    y = "Proportion"
  )

ggsave(paste0(pl_path, "barplot_celltype_prop_by_sample.png"), width = 10, height = 6, dpi = 300)
```

# Boxplots

## Cell type proportions stratified by condition

```{r boxplot-chunk, eval = !params$partial_data}
# This chunk will not run if partial_data == FALSE
prop_per_sample <- imap_dfr(spe_list, function(spe, name) {
  cd <- as.data.table(SummarizedExperiment::colData(spe))
  cd |>
    filter(!is.na(.data[[params$labels_key]])) |>
    dplyr::count(.data[[params$labels_key]], name = "count") |>
    mutate(proportion  = count / sum(count),
           sample      = name,
           tissue_type = paste(unique(as.character(cd$tissue_type)), collapse = ", ")) |> 
    dplyr::rename(cell_type = all_of(params$labels_key))
})

ggplot(prop_per_sample, aes(x = tissue_type, y = proportion, fill = tissue_type)) +
  geom_boxplot(outlier.shape = 21) +
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1b9e77", "c26" = "#d95f02")) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribuzione proporzioni cell_type per tissue_type",
       x = "Tissue type",
       y = "Proporzione")

ggsave(paste0(pl_path, "boxplot_celltype_prop_by_cond.png"), width = 10, height = 6, dpi = 300)
```

# ------------------------------------------------------------------------------

# Tabelle Risultati

## Tabelle proporzioni

Per campione
```{r}
celltype_table <- imap_dfr(spe_list, function(spe, name) {
        as.data.table(colData(spe)) |> 
          filter(!is.na(.data[[params$labels_key]])) |>
          dplyr::count(.data[[params$labels_key]], name = "count") |>
          rename(cell_type = all_of(params$labels_key)) |>
          mutate(proportion = count / sum(count),
                 sample = name)
})

celltype_matrix <- celltype_table |>
  select(sample, cell_type, proportion) |>
  pivot_wider(names_from = cell_type, values_from = proportion, values_fill = 0)

write_xlsx(celltype_matrix, paste0(res_path, "celltype_proportions_per_sample.xlsx"))
```


Per tipo di campione
```{r}
props_wide <- imap_dfr(spe_list, function(spe, name) {
  data.table(
    tissue_type = rep(as.character(unique(spe$tissue_type)),
                      length.out = length(colData(spe)[[params$labels_key]])),
    cell_type   = as.character(colData(spe)[[params$labels_key]])
  )
}) |>
  dplyr::count(tissue_type, cell_type, name = "n") |>
  group_by(tissue_type) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  select(tissue_type, cell_type, prop) |>
  pivot_wider(names_from = cell_type, values_from = prop, values_fill = 0)

 write_xlsx(props_wide, path = paste0(res_path, "celltype_by_tissue_type.xlsx"))
```


# Diagnostics

## Heatmap 

```{r ann-load-chunk, eval = params$diagnostics}
ann_list <- readRDS(paste0(Rd_path, params$ann_key))
```


```{r heatmap-plot-chunk, eval = params$diagnostics}
plots <- imap(ann_list, function(ann_el, name) {plotScoreHeatmap(ann_el, show.labels = FALSE)})

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"spat_annotation/diagnostic/heatmap/", name, "nolegend.png"), plot = plot, dpi = 700, create.dir = TRUE)
})
```

## Delta distribution

```{r delta-distrib-plot-chunk, eval = params$diagnostics}

plots <- imap(ann_list, function(ann_el, name) {plotDeltaDistribution(ann_el)})

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"spat_t_annotation/diagnostic/delta_distribution/", name, ".png"), plot = plot, dpi = 700, create.dir = TRUE)
})

```

### Markers heatmap per label

```{r eval=FALSE, include=FALSE}
plotMarkerHeatmap(pred.ann_will, spe_list$blocco1_c26STAT3, "Myonuclei_IIx_IIb")
```

```{r eval=FALSE, include=FALSE}
#unique(sce_ref$subclusters)
mionuclei <- c("Myonuclei_IIb", "Myonuclei_IIx", "Myonuclei_IIx_IIb", "Myonuclei_MTJ", "Myonuclei_IIx_IIa", "Myonuclei_NMJ", "Myonuclei_Trim63")

plotMarkerHeatmap(pred.ann_will, spe_list$blocco1_c26STAT3, label = "Myonuclei_IIx_IIb",
                  other.labels = mionuclei, top = 20, num.threads = 4)
```