---
title: "Data reading"
author: "Valerio Reffo"
date: "2025-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(jsonlite)
library(purrr)
library(dplyr)
#library(scuttle)
library(SpatialExperiment)
#library(SpatialFeatureExperiment)
library(sf)
library(sfarrow)
# to read spatialdata objects
library(ggnewscale)
library(SpatialData)
library(SingleCellExperiment)
library(anndataR)
source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
```

Paths
```{r}
#pl_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/cell_dec/"
Rd_path = "/mnt/europa/valerio/data/Rdata/"
zarr_path = "/mnt/europa/valerio/data/zarr_store/samples/"
#ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
#res_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/results/DEC_ANN/"
poly_path = "/mnt/europa/valerio/data/zarr_store/samples/"
sr_path = "/mnt/europa/valerio/data/zarr_store/adatas/spaceranger_tables/"
adata_path = "/mnt/europa/valerio/data/zarr_store/adatas/samples_adata/"
adata_arivis_path = "/mnt/europa/valerio/data/zarr_store/adatas/arivis_segmentation_tables/"
adata_016um_path = "/mnt/europa/valerio/data/zarr_store/adatas/binned_adatas/version_2.0.0/016um/"
```

# Concatenated samples

Al momento non è disponibile un file concatenato perché non riesco a farlo funzionare in python. Si può fare da R molto più tranquillamente.
```{r concat_data}
#(sdata <- readSpatialData("/mnt/europa/valerio/data/zarr_store/concat_all_samples_almost.zarr", anndataR=TRUE))
```

# List of all samples

This dictionary has all the info needed to read the data
```{r samples_dict, include=FALSE}
samples_dict <- fromJSON("/mnt/europa/valerio/repositories/cachetic_visiumHD/json/blocco_sample_bbox_dict.json")
```

Vector with samples name
```{r }
samples_key <- lapply(samples_dict, function(blocco) {
  lapply(blocco, function(sample) {
    return(sample$sample_key)
  })
}) |> unlist(use.names = FALSE)
samples_key
```

```{r reading_function, include=FALSE}
read_samples_list <- function(samples)
{
  spe_list <- lapply(samples, function(sample) {
    sdata <- readSpatialData(paste0(zarr_path, "/", sample, ".zarr"), anndataR=TRUE)
    #tables(sdata)$final_table
  })
  names(spe_list) = samples
  return(spe_list)
}
```

```{r samples_list}
sdata_list = read_samples_list(samples_key)
sdata_list
```

Per i dati in shared 
```{r}
read_shared_samples_list <- function(samples)
{
  spe_list <- map(samples, function(sample) {
    sdata <- readSpatialData(paste0(zarr_path, sample), anndataR=TRUE)
    #tables(sdata)$final_table
  })
  names(spe_list) = samples
  return(spe_list)
}
```

```{r}
sr_list = read_shared_samples_list(samples_key)
```

# ------------------------------------------------------------------------------

## Preparation of the different objects

## Single Cell Experiment to spatial experiment

```{r}
read_samples_adatas <- function(samples, data_path = "/mnt/europa/valerio/data/zarr_store/adatas/samples_adata/")
{
  spe_list <- map(samples, function(sample) {
    h5ad_path <- paste0(data_path, sample, ".h5ad")
    sce <- read_h5ad(h5ad_path, as = "SingleCellExperiment")
    blocco <- strsplit(sample, "_")[[1]][1]
    sampletype <- strsplit(sample, "_")[[1]][2]
    sce$blocco <- blocco
    sce$sampletype <- sampletype
    sce$ncounts <- colSums(assay(sce))
    sce$ngenes <- Matrix::colSums(assay(sce) > 0)
    sce$sample_id <- sample
    # add spatialcoordinate to the coordinate 
    spe <- toSpatialExperiment(sce, spatialCoordsNames = c("x_coord", "y_coord"))
    spe
  })
  names(spe_list) = samples
  return(spe_list)
}
```

```{r}
samples_key_no2 <- samples_key[!grepl("blocco2", samples_key)]
samples_key_b2 <- samples_key[grepl("blocco2", samples_key)]
spe_list = read_samples_adatas(samples_key_b2, data_path = adata_016um_path)
#sce_list
```

Filtering low count nuclei
```{r}

# for(nm in names(sce_list)) {
#   sce_list[[nm]] <- sce_list[[nm]][, colData(sce_list[[nm]])$ncounts >= quantile(colData(sce_list[[nm]])$ncounts, 0.05) &
#                        colData(sce_list[[nm]])$ngenes >= quantile(colData(sce_list[[nm]])$ngenes, 0.05)]
# }
```

## Single sce object

```{r}
sce_merged <- do.call(cbind, sce_list)
sce_merged
```

Ok le dimensioni corrispondono

```{r}
sum(sapply(sce_list, ncol))
dim(sce_merged)
```


## Pseudobulk

```{r}
sce2pb <- \(sce_list) {
  pb_list <- lapply(sce_list, function(sce) {
  scuttle::aggregateAcrossCells(sce, ids = colData(sce)$tissue_type, use.assay.type = "X")
  })
  all_counts <- do.call(cbind, lapply(pb_list, assay))
  all_colData <- do.call(rbind, lapply(pb_list, colData))
  final <- SingleCellExperiment(assays=list(counts=all_counts), colData=all_colData)
  return(final)
}
```

```{r}
pb <- sce2pb(sce_list)
```

### Psuedobulk dalla lista di spe

Procedo a rifare i pb senza blocco 7 e filtrando i "to_discard" e raggruppato per cell type

```{r}
# 1. Filter the list to remove samples starting with "blocco7"
# grep(..., invert = TRUE) returns indices/names that DO NOT match the pattern
spe_list_clean <- spe_list[grep("^blocco7", names(spe_list), invert = TRUE)]

# 2. Iterate over the remaining list to filter spots where to_discard is TRUE
spe_list_clean <- lapply(spe_list_clean, function(spe) {
  # Keep columns where to_discard is FALSE (or NA, usually safe to treat NA as FALSE here or handle explicitly)
  spe[, !spe$to_discard]
})

# 3. Create Pseudobulk object
# First, we usually need to combine the list into a single SingleCellExperiment/SpatialExperiment
# to perform aggregation across the whole dataset efficiently.
spe_combined <- do.call(cbind, spe_list_clean)

# 4. Aggregate (Pseudobulk) by "labels_singler" and "sample_id"
# We usually group by Sample AND Cell Type to get pseudobulk profiles per sample-celltype pair.
# If you only group by 'labels_singler', you will average across ALL samples (which is rarely the goal).
# Assuming 'sample_id' is the column identifying the sample name in colData.

# Define the grouping factors
ids <- colData(spe_combined)[, c("labels_singler", "sample_id")]

# Perform the aggregation (summing counts is standard for pseudobulk)
pb_spe <- aggregateAcrossCells(spe_combined, ids = ids)
```


## Spatial data

Aggiungo spatial coordinates ai sce

```{r}
spatialc <- read.csv("/mnt/europa/valerio/data/data_tables/spatial_coords.csv", sep=",", header = FALSE)
colnames(spatialc) <- c("y", "x")
spatialc$sample_id <- sce_merged$sample_id
head(spatialc)
dim(spatialc)
dim(sce_merged)
dim(spatialc)[1] == dim(sce_merged)[2]
```

### Spatial experiment merged
```{r}
sce_merged$x <- spatialc$x
sce_merged$y <- spatialc$y
(spe_merged <- toSpatialExperiment(sce_merged,
    spatialCoordsNames = c("y", "x")))
```

### Spatial experiment list

```{r}
add_spatial_coords <- function(sce_list, spatial_coords, sample_col = "sample_id") {
  lapply(sce_list, function(sce) {
    # Identify the sample id for this SCE
    sample_id <- unique(colData(sce)[[sample_col]])
    # Subset spatialc for this sample
    spatial_sub <- spatial_coords[spatial_coords[[sample_col]] == sample_id, ]
    # Add x and y to colData
    colData(sce)$x <- spatial_sub$x
    colData(sce)$y <- spatial_sub$y
    return(sce)
  })
}
```

Let's add spatial coordinates to every element of sce_list
```{r}
sce_list <- add_spatial_coords(sce_list, spatialc, sample_col = "sample_id")
sapply(sce_list, function(sce) all(c("x", "y") %in% colnames(colData(sce))))
```

Convert to spe
```{r}
spe_list <- lapply(sce_list, function(sce) {
  toSpatialExperiment(sce, spatialCoordsNames = c("y", "x"))
})
```


## Spe grouped by sampletypes

```{r}
# Get the unique sample types
sampletypes <- unique(spe_merged$sampletype)

# Create a list with one SpatialExperiment per sample type
spe_st <- lapply(sampletypes, function(st) {
  spe_merged[, colData(spe_merged)$sampletype == st]
})

# Optionally, name the list
names(spe_st) <- sampletypes
```

# spe_merged without blocco7

Devo togliere b7 e aggiungere il vettore cell_type e tissue_type agli spe_merged e poi ricreare lo pseudo bulk. Inoltre anche in_tissue così i grafici non stressano

```{r}
# 1) Costruisci la tabella mapping da spe_list
mapping_df <- imap_dfr(spe_list_wb7, function(spe, name) {
  cd <- as.data.frame(colData(spe))
  # assicurati che le colonne esistano e siano character (evita factor surprises)
  if (!("cell_id" %in% names(cd))) stop("In spe_list[['", name, "']] manca la colonna 'cell_id'")
  if (!("cell_type" %in% names(cd))) stop("In spe_list[['", name, "']] manca la colonna 'cell_type'")

  tibble(
    sample_id = name,
    cell_id   = as.character(cd$cell_id),
    cell_type = as.character(cd$cell_type)
  )
})

# 2) Controlli rapidi: duplicati o mismatch
dups <- mapping_df |> 
  count(sample_id, cell_id)  |> 
  filter(n > 1)

if (nrow(dups) > 0) {
  stop("Trovati duplicati in mapping_df per sample_id+cell_id: controlla i seguenti (esempi):\n",
       paste(utils::head(paste0(dups$sample_id, "/", dups$cell_id), 10), collapse = ", "))
}

# 3) Prepara il colData di spe_merged come data.frame

# 4) Left join per ottenere cell_type allineato all'ordine di spe_merged
cd_joined <- as.data.frame(colData(spe_merged))  |> 
  mutate(sample_id = as.character(sample_id),
         cell_id   = as.character(cell_id)) |> 
  left_join(mapping_df, by = c("sample_id", "cell_id"))

# 5) Verifiche: tutte le righe hanno un valore corrispondente?
missing_n <- sum(is.na(cd_joined$cell_type))
if (missing_n > 0) {
  warning(sprintf("Attenzione: %d colonne in spe_merged non hanno corrispondenza in mapping_df (cell_type NA).",
                  missing_n))
  # se vuoi fermare in caso di mismatch, usa stop() invece di warning()
}

# 6) Assegna in un'unica operazione al colData
colData(spe_merged)$cell_type <- cd_joined$cell_type

# 7) (Opzionale) converti in factor con livelli coerenti se vuoi
all_levels <- unique(na.omit(mapping_df$cell_type))
colData(spe_merged)$cell_type <- factor(colData(spe_merged)$cell_type, levels = all_levels)

spe_merged$tissue_type <- spe_merged$sampletype
spe_merged$sampletype <- NULL
spe_merged$in_tissue <- TRUE
spe_merged_wb7 <- spe_merged[, spe_merged$blocco %nin% "blocco7"]
```

# spe_list without blocco 7 and coldata corrected

```{r}
spe_list_wb7 <- spe_list[!startsWith(names(spe_list), "blocco7_")]
```

Cambio il nome della colonna sampletype in tissue_type, cosi siamo uguali a emma
e aggiungo 
```{r}
spe_list_wb7 <- imap(spe_list_wb7, function(spe, name) {
  spe$tissue_type = spe$sampletype
  spe$sampletype = NULL
  spe$in_tissue <-  TRUE
  spe
})
```

```{r}
spe_list <- imap(spe_list, function(spe, name) {
  spe$tissue_type = spe$sampletype
  spe$sampletype = NULL
  spe$in_tissue <-  TRUE
  spe
})
```


# Pseudobulk without blocco 7

Creiamo oggetto pseudobulk per i vari cell_type
```{r}
# Create pseudo-bulk for each cell_type in each sample
pb_ct_wb7 <- aggregateAcrossCells(spe_merged_wb7, use.assay.type = "X",
                                id=DataFrame(label = spe_merged_wb7$cell_type,
                                             sample = spe_merged_wb7$sample_id))
pb_ct_wb7  <- logNormCounts(pb_ct_wb7,  assay.type = "X") #altre normalizzazioni?
(colnames(pb_ct_wb7) <- paste(pb_ct_wb7$cell_type, pb_ct_wb7$sample_id, sep = "_"))
```

Rifaccio pseudobulk per tissue_type (aka sampletype) e tolgo blocco 7
```{r}
pb_tt_wb7 <- aggregateAcrossCells(spe_merged_wb7, use.assay.type = "X",
                                id=DataFrame(label = spe_merged_wb7$tissue_type,
                                             sample = spe_merged_wb7$sample_id))
pb_tt_wb7  <- logNormCounts(pb_tt_wb7,  assay.type = "X") #altre normalizzazioni?
(colnames(pb_tt_wb7) <- pb_tt_wb7$sample_id)
```

# rifaccio gli oggetti spe_merged e pb con l'annotazione SingleR

```{r}
load(paste0(Rd_path, 'spe_list_nuclei_ann.RData'))
```

## Spe merged
```{r}
spe_merged_sr <- do.call(cbind, spe_list_up)
```

without blocco 7
```{r}
spe_list_up_wb7 <- spe_list_up[!startsWith(names(spe_list_up), "blocco7_")]
spe_merged_sr_wb7 <- do.call(cbind, spe_list_up_wb7)
```

## Pseudobulk SingleR

Creiamo oggetto pseudobulk per i vari cell_type
```{r}
pb_ct <- aggregateAcrossCells(spe_merged_sr, use.assay.type = "X",
                                id=DataFrame(label = spe_merged_sr$labels_singler,
                                             sample = spe_merged_sr$sample_id))
pb_ct  <- logNormCounts(pb_ct,  assay.type = "X") 
colnames(pb_ct) <- paste(pb_ct$labels_singler, pb_ct$sample_id, sep = "_")
```

without blocco 7
```{r}
pb_ct_wb7 <- aggregateAcrossCells(spe_merged_sr_wb7, use.assay.type = "X",
                                id=DataFrame(label = spe_merged_sr_wb7$labels_singler,
                                             sample = spe_merged_sr_wb7$sample_id))
pb_ct_wb7  <- logNormCounts(pb_ct_wb7,  assay.type = "X") 
colnames(pb_ct_wb7) <- paste(pb_ct_wb7$labels_singler, pb_ct_wb7$sample_id, sep = "_")
```

# Spaceranger segmentation data reading 

Select the first blocco
```{r}
#(samples_b1 <- samples_key[startsWith(samples_key, "blocco1_")])
samples_2 <- samples_key[stringr::str_detect(samples_key, "^blocco[12]")]
```


```{r reading_function, include=FALSE}
read_samples_list <- function(samples)
{
  spe_list <- map(samples, function(sample) {
    sdata <- readSpatialData(paste0("/mnt/europa/valerio/data/zarr_store/spaceranger_v4/no_cell_expans/samples/", sample), anndataR=TRUE)
    sdata
  })
  names(spe_list) = samples
  return(spe_list)
}
```

```{r samples_list}
sr_list = read_samples_list(samples_2)
sr_list
```

Extract tables and create spatial experiment
```{r}
sr_spe_list <- imap(sr_list, function(sdata, name){
  sce <-  sdata$tables$segmentation_counts
  spe <- toSpatialExperiment(sce, spatialCoordsNames = c("x", "y"))
  # rename assay counts
  assayNames(spe)[assayNames(spe) == "X"] <- "counts"
  # filter nuclei with zero counts
  spe <- spe[, colSums(counts(spe)) > 0]
  spe
})
sr_spe_list

```

## Spaceranger reading from adata

```{r}
read_adatas_list <- function(samples, data_path = NULL)
{
  adata_list <- map(samples, function(sample) {
    adata <- read_h5ad(paste0(data_path, sample, ".h5ad"),   as = "SingleCellExperiment")
    adata
  })
  names(adata_list) = samples
  return(adata_list)
}
```

```{r}
sr_list = read_adatas_list(samples_key)
sr_list
```
Correct the sr_list to become spatialExperiments
```{r}
sr_spe_list <- imap(sr_list, function(sce, name){
  spe <- toSpatialExperiment(sce, spatialCoordsNames = c("x", "y"))
  # rename assay counts
  assayNames(spe)[assayNames(spe) == "X"] <- "counts"
  # filter nuclei with zero counts
  spe <- spe[, colSums(counts(spe)) > 0]
  spe
})
sr_spe_list
```

```{r}
read_adatas_list <- function(samples, data_path = NULL)
{
  adata_list <- map(samples, function(sample) {
    sce <- read_h5ad(paste0(data_path, sample, ".h5ad"),   as = "SingleCellExperiment")
    blocco <- strsplit(sample, "_")[[1]][1]
    sampletype <- strsplit(sample, "_")[[1]][2]
    sce$blocco <- blocco
    sce$sampletype <- sampletype
    sce$sample_id <- sample
    sce
  })
  names(adata_list) = samples
  return(adata_list)
}
```

```{r}
spe_list <- read_adatas_list(samples_key, data_path = adata_path)
```


# Add labels of RCTD doublet into the spe_list_up where all info are stored togheter

I need to add the column in the spe list but the blocco7 wasn't annotated so let's skip them. A for loop is more handly.

```{r}
spe_list_up<- mappingLabels(spe_list_up, ann_list_RCTD)
```

# Spatial Feature Experiment

Let's try to convert from SpatialExperiment a single sample
```{r}
# sfe_prova <- toSpatialFeatureExperiment(spe_prova)
# 
# # Read the nuclei, they are here:
# # "/mnt/europa/valerio/data/zarr_store/samples/blocco1_c26STAT3.zarr/shapes/filtered_nuclei/shapes.parquet"
# # Read the Parquet file into a dataframe/tibble
# poly <- st_read_parquet("/mnt/europa/valerio/data/zarr_store/samples/blocco1_c26STAT3.zarr/shapes/filtered_nuclei/shapes.parquet")
# poly$cell_id <- poly$index_left
# poly$index_left <- NULL
# poly
# 
# # filtering nuclei in the sample and reordering
# poly_filtered <- poly |>
#   filter(cell_id %in% sfe$cell_id) |>
#   slice(match(sfe$cell_id, cell_id))
# rownames(poly_filtered) <- colnames(sfe)
# 
# # add to the sfe
# # MARGIN: 1 = righe (geni)
# #         2 = colonne (nuclei)
# dimGeometry(sfe, "filtered_nuclei", MARGIN = 2) <- poly_filtered

# # try the fun
#sfe_prova <- add2sfe(sfe_prova, shape_name = "filtered_nuclei", MARGIN = 2)
```

Let's read the shapes for each sample

```{r}
poly_list <- read_shapes(poly_path, samples_key)
names(poly_list) <- samples_key
```

```{r}
sfe_prova <- toSpatialFeatureExperiment(spe_list$blocco1_c26STAT3) |> 
  add2sfe(poly_list$blocco1_c26STAT3, MARGIN = 2, )
```


All samples togheter
```{r}

sfe_list <- map2(spe_list, poly_list, function(spe, poly){
    sfe <- toSpatialFeatureExperiment(spe)
    sfe <- add2sfe(sfe, poly, MARGIN = 2, name = "filtered_nuclei")
    sfe
  })

```

# Saving objects

```{r}
save(pb_tt_wb7, file = paste0(Rd_path, "pseudobulk_tissuetype_wb7.RData"))
save(pb_ct_wb7, file = paste0(Rd_path, "pseudobulk_celltype_wb7.RData"))
save(sce_list, file = paste0(Rd_path, "sce_list_nuclei.RData"))
save(sce_merged, file = paste0(Rd_path, "sce_merged_nuclei.RData"))

save(spe_list, file = paste0(Rd_path, "spe_list_nuclei.RData"))
save(spe_list_wb7, file = paste0(Rd_path, "spe_list_nuclei_dec_wb7.RData"))
save(spe_merged, file = paste0(Rd_path, "spe_merged_nuclei.RData"))
save(spe_merged_wb7, file = paste0(Rd_path, "spe_merged_nuclei_wb7.RData"))
# oggetti risalvati con annotazione singleR
save(spe_merged_sr, file = paste0(Rd_path, "spe_merged_nuclei_sr.RData"))
save(spe_merged_sr_wb7, file = paste0(Rd_path, "spe_merged_nuclei_sr_wb7.RData"))
saveRDS(pb_ct, file = paste0(Rd_path, "pseudobulk_celltype_SingleR"))
save(pb_ct_wb7, file = paste0(Rd_path, "pseudobulk_celltype_SingleR_wb7.RData"))
# Spatial Features Experiment (spe with nuclei polygons)
saveRDS(sfe_list, file = paste0(Rd_path, "sfe_list_annotated"))
# pseudobulk wb7 filtrata per to_discard
saveRDS(pb_spe, file = paste0(Rd_path, "pseudobulk_celltype_SingleR_wb7_wdiscard"))

# oggetto aggiornato con tutto (3/11/25)
saveRDS(spe_list_up, file = paste0(Rd_path, "spe_list_nuclei_ann"))
saveRDS(spe_list_12, file = paste0(Rd_path, "spe_list_nuclei_ann_b1_b2")) 

# spaceranger - blocco1
saveRDS(sr_spe_list, file = paste0(Rd_path, "spaceranger_b1_spe"))
# spaceranger - blocco1 e blocco2 no cell expansion
saveRDS(sr_spe_list, file = paste0(Rd_path, "spaceranger_spe_nocell_expans"))
# spaceranger - blocco 1 e blocco2 cell expansion 2 microns
saveRDS(sr_spe_list, file = paste0(Rd_path, "spaceranger_spe_cell_expans_2um"))

# arivis segmentation - e 016um new pipelined
saveRDS(spe_list, file = paste0(Rd_path, "spe_list_arivis_segmentation_no_b2.rds"))
saveRDS(spe_list_all, file = paste0(Rd_path, "spe_list_arivis_segmentation.rds"))
saveRDS(spe_list, file = paste0(Rd_path, "spe_list_016um_b2.rds"))
```




```{r}
#rm(spe_list_up)

spe_list <- readRDS(paste0(Rd_path, "spe_list_nuclei_ann"))
spe_list_12 <- spe_list[stringr::str_detect(names(spe_list), "^blocco[12]")]
```

