---
title: "nuclei_filtering"
output: html_document
params:
  labels_key: "labels_singler"
  #spe_key: "spaceranger_spe_nocell_expans"
  spe_key: "spe_list_nuclei_ann"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(ggplot2)
library(ggspavis)
library(ggnewscale)
library(scuttle)
library(SpatialExperiment)
library(data.table)
library(SingleR)
library(dplyr)
library(tidyr)
library(SingleCellExperiment)
library(SpaceTrooper)
library(patchwork)
source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
```

Paths
```{r}
pl_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/filtered_nuclei/"
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/results/"
```

```{r}
color_vector <- readRDS(paste0(Rd_path, "color_cell_dec"))
spe_list <- readRDS(paste0(Rd_path, params$spe_key))
```

# Single sample analysis

```{r}
colData(spe_list_up$blocco1_c26STAT3)

spe_prova <- spe_list_up$blocco3_c26murf1
spe_prova <-  vHD_spatialPerCellQC(spe_prova)
```

Plots of QC metrics

```{r}
plotMetricHist(spe_prova, metric="sum") # total counts
plotMetricHist(spe_prova, metric="Area_um", bins = 50) # area nuclei in micron
plotMetricHist(spe_prova, metric="area", bins = 50) # are non in micron
plotMetricHist(spe_prova, metric="CountArea", bins = 50) # rapporto sum/area_um
plotMetricHist(spe_prova, metric="log2CountArea", bins = 50) # log2
plotMetricHist(spe_prova, metric="AspectRatio", bins = 50) # major axis length / minor axis length
```


Identify spatial outliers based on cell area (Area_um) and log2CountArea 

```{r}

spe_prova <- computeSpatialOutlier(spe_prova, computeBy="Area_um", method="mc")
spe_prova <- computeSpatialOutlier(spe_prova, computeBy="log2CountArea", method="mc")
#spe_prova <- computeSpatialOutlier(spe_prova, computeBy="log2AspectRatio", method="mc")
#spe_prova <- computeSpatialOutlier(spe_prova, computeBy="sum", method="mc")

names(colData(spe_prova))
base::table(Area_um = spe_prova$Area_um_outlier_mc)
base::table(CountArea = spe_prova$log2CountArea_outlier_mc)
base::table(spe_prova$log2AspectRatio_outlier_mc)
base::table(Area_outlier = spe_prova$Area_um_outlier_mc, log2countArea_outlier = spe_prova$log2CountArea_outlier_mc,
            log2AR = spe_prova$log2AspectRatio_outlier_mc)
```

To retrieve the thresholds (in the attr of the coldata column)

```{r}
attr(spe_prova$log2CountArea_outlier_mc, "thresholds")
```

```{r}
plotMetricHist(spe_prova, metric="Area_um", useFences="Area_um_outlier_mc")
#plotMetricHist(spe_prova, metric="sum", useFences="sum_outlier_mc")
plotMetricHist(spe_prova, metric="log2CountArea", useFences="log2CountArea_outlier_mc")
plotMetricHist(spe_prova, metric="CountArea")
```

Rimuovo le cellule con aree grandi (HIGH = 632) e con rapporto tra conteggio e area basso (344)
ma prima visualizziamo meglio questi nuclei prima di rimuoverli

```{r}
plotCentroids(spe_prova, colourBy = "Area_um_outlier_mc")
plotCentroids(spe_prova, colourBy = "log2CountArea_outlier_mc")
```

I nuclei con basso log2CountArea sono per lo più esterni, quindi rimuovo senza problemi, mentre i nuclei troppo grandi sembrano sparsi abbastanza casualmente nel tessuto. rimuovo anche questi. 

```{r}
table(spe_prova$log2CountArea_outlier_mc == "LOW")
table(spe_prova$Area_um_outlier_mc == "HIGH")
table(spe_prova$Area_um_outlier_mc == "HIGH" | spe_prova$log2CountArea_outlier_mc == "LOW")
spe_prova <- spe_prova[, !(spe_prova$Area_um_outlier_mc == "HIGH" | spe_prova$log2CountArea_outlier_mc == "LOW")]

```


Cell type dopo aver eliminato gli outliers

```{r}
plotCoords(spe_prova, annotate = "labels_singler",
             point_size = 0.05,
             x_coord = "x_um", y_coord = "y_um") +
    scale_color_manual(values = color_vector) +
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    )
```

```{r}
plotCoords(spe_prova, annotate = "labels_singler", point_size = 0.05)
```



--------------------------------------------------------------------------------
# All samples analysis 

Calculate QC metrics for every sample and put in the colData and the coordinate in micron instead of pixels
```{r}
spe_list_up <- imap(spe_list, \(spe, name) {vHD_spatialPerCellQC(spe)})
```

Plot and save most important metrics

```{r}
metric_to_plot = "log2CountArea"
plots <- imap(spe_list_up, \(spe, name){
  p <- plotMetricHist(spe, metric=metric_to_plot, bins = 50) +
          ggtitle(paste0(name))
  p
})

p1 <- wrap_plots(plots, ncol = 2) + 
  plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt1'))
# p2 <- wrap_plots(plots[c(13:24)], ncol = 4) + 
#   plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt2'))

ggsave(paste0(pl_path,"metrics/", metric_to_plot, "_part1.png"), plot = p1, dpi = 300, units = "cm", width = 50, height = 30)
ggsave(paste0(pl_path,"metrics/", metric_to_plot, "_part2.png"), plot = p2, dpi = 300, units = "cm", width = 50, height = 30)
```

```{r}
# plots <- plot_metrics(spe_list_up, metric_to_plot = "sum", save = TRUE, bin = 50)
# non funziona ma non ho voglia di farla andare
```

I'll look only at the area of nuclei and the ratio between counts and area of the nuclei
Since there is only one sample that has log2CountArea asymmetric enough, I'll use sc method to determine the outliers in that metric

```{r}
spe_list_up <- imap(spe_list_up, \(spe, name){
  spe <- computeSpatialOutlier(spe, computeBy="Area_um", method="mc") # "mc" for asymmetric distribution of the choosen metric
  spe <- computeSpatialOutlier(spe, computeBy="log2CountArea", method="sc") # "sc" for symmetric distribution of the choosen metric
  spe
})
```

Let's investigate the fences for area_um 

```{r}
spe_list_up |> 
  map(function(spe) {
    attr(spe$Area_um_outlier_mc, 'thresholds')
  })
```


```{r}
outlier_summary_df <- imap_dfr(spe_list_up, \(spe, name) {
  
  colData(spe) |>
    as.data.frame() |>
    select(Area_um_outlier_mc, log2CountArea_outlier_sc) |>
    mutate(across(everything(), as.character)) |>
    pivot_longer(
      cols = everything(),
      names_to = "metric",
      values_to = "outlier_type"
    ) |>
    count(metric, outlier_type, name = "count") |>
    mutate(sample = name)
})

```

```{r}
outlier_summary_df |> 
  filter(
    metric == "Area_um_outlier_mc" & outlier_type %in% c("HIGH","LOW")
  ) |> 
  select(sample, count, outlier_type, metric)


outlier_summary_df |> 
  filter(
    metric == "log2CountArea_outlier_sc" & outlier_type == "LOW"
  ) |> 
  select(sample, count, outlier_type, metric)


```

Plot and save most important metrics with the outliers thresholds

```{r}
metric_to_plot = "Area_um"
plots <- imap(spe_list_up, \(spe, name){
  p <- plotMetricHist(spe, metric=metric_to_plot, bins = 50, useFences= paste0(metric_to_plot,"_outlier_mc")) +
          labs(color="Fences outlier") +
          ggtitle(name)
  p
})

p1 <- wrap_plots(plots, ncol = 4) + 
  plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt1'))
# p2 <- wrap_plots(plots[c(13:24)], ncol = 4) + 
#   plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt2'))
 
ggsave(paste0(pl_path,"outliers/", metric_to_plot, "_part1.png"), plot = p1, dpi = 300, units = "cm", width = 50, height = 30)
ggsave(paste0(pl_path,"outliers/", metric_to_plot, "_part2.png"), plot = p2, dpi = 300, units = "cm", width = 50, height = 30)
```

Plots of the nuclei centroids colored by outliers info

```{r}

# "Area_um_outlier_mc"
# "log2CountArea_outlier_sc"


metric_to_plot = "log2CountArea_outlier_sc"
plots <- imap(spe_list_up, \(spe, name){
  p <- plotCentroids(spe, colourBy = metric_to_plot, size = 0.5) +
          ggtitle(name)
  p
})

# p1 <- wrap_plots(plots[c(1:12)], ncol = 4) + 
#   plot_annotation(title = paste0('Nuclei outliers - ', metric_to_plot))
# p2 <- wrap_plots(plots[c(13:24)], ncol = 4) + 
#   plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt2'))
# 
# ggsave(paste0(pl_path,"area_outlier/", metric_to_plot, "_part1.png"), plot = p1, dpi = 300, units = "cm", width = 50, height = 30)
# ggsave(paste0(pl_path,"area_outlier/", metric_to_plot, "_part2.png"), plot = p2, dpi = 300, units = "cm", width = 50, height = 30)

iwalk(plots, function(plot, name){
 ggsave(paste0(pl_path,"CountArea_outlier/", name, ".png"), plot = plot, dpi = 300, units = "cm", width = 50, height = 30)
  #ggsave(paste0(pl_path,"area_outlier/", name, ".png"), plot = plot, dpi = 300, units = "cm", width = 50, height = 30)
})
```

# Filtering nuclei

## Area filtering

It's difficult to define a standard measure for myonucleus, so I decided arbitrarly that the minimum will be 5 mu squared and the maximum will be 100 mu squared. We are comfortably larger and smaller, because it can happen to have a tip of a nucleus that it will look smaller or an out of focus big nucleus that will look bigger. so I apply this thresholds to every samples
```{r}
nuclei_filtering_fun <- function(spatial_exp, 
                                 feature = 'Area_um', 
                                 min_value = 0, 
                                 max_value = 150,
                                 keep_na = FALSE) {
  # Basic checks
  if (!inherits(spatial_exp, c("SpatialExperiment", "SingleCellExperiment", "SummarizedExperiment"))) {
    stop("`spatial_exp` must be a SpatialExperiment / SingleCellExperiment (or SummarizedExperiment-derived) object.")
  }

  cd <- SummarizedExperiment::colData(spatial_exp)
  if (!(feature %in% colnames(cd))) {
    stop(sprintf("Column '%s' not found in colData(spatial_exp). Available columns: %s",
                 feature, paste(colnames(cd), collapse = ", ")))
  }
  # Coerce to numeric (NA where coercion fails)
  features_vals <- suppressWarnings(as.numeric(cd[[feature]]))
  
  # Build the logical mask
  if (keep_na) {
    keep <- is.na(features_vals) | (features_vals >= min_value & features_vals <= max_value)
  } else {
    keep <- !is.na(features_vals) & (features_vals >= min_value & features_vals <= max_value)
  }
  
  n_before <- dim(spatial_exp)[2]
  n_after <- sum(keep)
  message(sprintf("Filtering by %s: %d -> %d observations (kept NAs: %s).",
                  feature, n_before, n_after, ifelse(keep_na, "TRUE", "FALSE")))

  # Subset columns (observations / cells / spots are columns in SingleCellExperiment)
  filtered_spatial_exp <- spatial_exp[, keep, drop = FALSE]

  return(filtered_spatial_exp)
}
```

And then the actual filtering
```{r}
filtered_spe <- spe_list_up |> 
  map(nuclei_filtering_fun(min_value = 0, max_value = 150))
```

## log2CoutArea filtering

Here I'll follow the thresholds defined by the methodology, medcouple if asymmetrics and MAD for symmetrics distribution of feature value

```{r}
filtered_spe <- filtered_spe |> 
  map(function(spe) {
    keep <- spe$log2CountArea_outlier_sc == "NO"
    spe <- spe[, keep]
  })
```

# Save the new object

```{r}
saveRDS(filtered_spe, file = paste0(Rd_path, "filtered_spe_annotated_28Nov"))
```


# Results of the filtering

Remove the to_discard nuclei to have only the good ones
```{r}
filtered_spe <- filtered_spe |> 
  map(function(spe) {
    spe <- spe[, !spe$to_discard]
    spe
  })
```


## Plots 

Barplots of cell type proportions
```{r}
prop_df_ordered <- imap_dfr(filtered_spe, function(spe, name) {
  tab <- base::table(colData(spe)[[params$labels_key]])
  prop <- prop.table(tab)
  tissue_type <- unique(spe$tissue_type)
  data.table(
    sample_id = name,
    cell_type = names(prop),
    proportion = as.numeric(prop),
    tissue_type = as.character(tissue_type)
  )
}) |> 
  # 1. First, arrange the entire dataframe by tissue_type, then by sample_id
  arrange(tissue_type, sample_id) |>
  # 2. Then, create a factor from sample_id that follows its current order
  mutate(sample_id = fct_inorder(sample_id))

# vertical bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +                             # geom_col uses the y value directly (identity)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +# use your named color_vector; otherwise use scale_fill_brewer()
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

```{r}
prop_df_condition <- imap_dfr(filtered_spe, function(spe, name) {
  data.table(
    tissue_type = as.character(spe$tissue_type),
    cell_type = colData(spe)[[params$labels_key]]
  ) |> 
    filter(!is.na(cell_type))
}) |> 
  arrange(tissue_type) |> 
  dplyr::count(tissue_type, cell_type) |>
  group_by(tissue_type) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()


ggplot(prop_df_condition, aes(x = tissue_type, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per condizione",
    x = "Condizione (tissue_type)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


Boxplots - Cell type proportions stratified by condition

```{r boxplot-chunk, eval = !params$partial_data}
# This chunk will not run if partial_data == FALSE
prop_per_sample <- imap_dfr(filtered_spe, function(spe, name) {
  cd <- as.data.table(SummarizedExperiment::colData(spe))
  cd |>
    filter(!is.na(.data[[params$labels_key]])) |>
    dplyr::count(.data[[params$labels_key]], name = "count") |>
    mutate(proportion  = count / sum(count),
           sample      = name,
           tissue_type = paste(unique(as.character(cd$tissue_type)), collapse = ", ")) |> 
    dplyr::rename(cell_type = all_of(params$labels_key))
})

ggplot(prop_per_sample, aes(x = tissue_type, y = proportion, fill = tissue_type)) +
  geom_boxplot(outlier.shape = 21) +
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1b9e77", "c26" = "#d95f02")) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribuzione proporzioni cell_type per tissue_type",
       x = "Tissue type",
       y = "Proporzione")

ggsave(paste0(pl_path, "boxplot_celltype_prop_by_cond.png"), width = 10, height = 6, dpi = 300)
```
