---
title: "From python to R data reading"
author: "Valerio Reffo"
date: "2025-11-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(jsonlite)
library(anndataR)
library(SpatialExperiment)
library(dplyr)
library(purrr)
# to read spatialdata objects
library(ggnewscale)
library(SpatialData)
library(SingleCellExperiment)
```

Paths
```{r}
#pl_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/cell_dec/"
#Rd_path = "/mnt/europa/valerio/data/Rdata/"
zarr_path = "/mnt/europa/shared/sandri_visiumHD_data/bins/"
#ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
#res_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/results/DEC_ANN/"
```

# List of all samples

This dictionary has all the info needed to read the data

```{r samples_dict, include=FALSE}
samples_dict <- fromJSON("/mnt/europa/shared/sandri_visiumHD_data//blocco_sample_bbox_dict.json")
```

Vector with samples name

```{r }
samples_key <- lapply(samples_dict, function(blocco) {
  lapply(blocco, function(sample) {
    return(sample$sample_key)
  })
}) |> unlist(use.names = FALSE)
samples_key
```

Function to read the data and put them in a list, you can add steps to clean up the data if needed - remember that you cannot overwrite a zarr store, so save them back in R format and put them in another folder of the directory with appropriate naming (version_1.x.x) 

```{r reading_function, include=FALSE}
read_samples_list <- function(samples)
{
  spe_list <- map(samples, function(sample) {
    sdata <- readSpatialData(paste0(zarr_path, sample), anndataR=TRUE)
    sdata
  })
  names(spe_list) = samples
  return(spe_list)
}
```

If you want a subset 

```{r}
samples_b1b2 <- samples_key[stringr::str_detect(samples_key, "^blocco[12]")]
```

```{r}
sr_list = read_samples_list(samples_key)
sr_list
```

Extract tables and create spatial experiment
```{r}
sr_spe_list <- imap(sr_list, function(sdata, name){
  sce <-  sdata$tables$filtered
  spe <- toSpatialExperiment(sce, spatialCoordsNames = c("y", "x"))
  # rename assay counts
  assayNames(spe)[assayNames(spe) == "X"] <- "counts"
  # filter nuclei with zero counts
  spe <- spe[, colSums(counts(spe)) > 0]
  spe
})
sr_spe_list
```