---
title: "spaceranger_filtering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(ggplot2)
library(ggspavis)
library(ggnewscale)
library(scuttle)
library(SpatialExperiment)
library(data.table)
library(SingleR)
library(dplyr)
library(tidyr)
library(SingleCellExperiment)
library(SpaceTrooper)
library(patchwork)
source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
```

Paths
```{r}
pl_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/filtered_nuclei/"
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/results/"
```

Data
```{r}
load(paste0(Rd_path, "spe_list_nuclei_ann.RData"))
load(paste0(Rd_path, "color_cell_dec.RData"))
#save(sr_spe_list, file = paste0(Rd_path, "spaceranger_b1_spe.RData"))
```

# QC metrics

To integrate the csv file with morphological info needed for the task

```{r}

morph_list <- list()
nuclei_morph <- read.csv("/mnt/europa/valerio/data/data_tables/b1_c26foxO_nuclei_morph_SpaceRanger.csv")
morph_list[["blocco1_c26foxO"]] <- nuclei_morph
```

```{r}
sr_spe_list <- map2(sr_spe_list, morph_list, function(spe, morf_df){
  
  col_data_df <- as.data.frame(colData(spe)) |> 
    tibble::rownames_to_column(var = "cell_id")
  
  merged_coldata <- left_join(col_data_df, morf_df, by = "cell_id")
  rownames(merged_coldata) <- merged_coldata$cell_id
  
  colData(spe) <- DataFrame(merged_coldata)
  spe
})
```


```{r}
sr_spe_list <- imap(sr_spe_list, \(spe, name) {vHD_spatialPerCellQC(spe)})
```

Plot and save most important metrics

```{r}
metric_to_plot = "AspectRatio"
plots <- imap(sr_spe_list, \(spe, name){
  p <- plotMetricHist(spe, metric=metric_to_plot, bins = 50) +
          ggtitle(paste0(name))
  p
})

p1 <- wrap_plots(plots, ncol = 3) + 
  plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt1'))
p1

ggsave(paste0(pl_path,"metrics/", metric_to_plot, "_part1.png"), plot = p1, dpi = 300, units = "cm", width = 50, height = 30)
```

I'll look only at the area of nuclei and the ratio between counts and area of the nuclei
Since there is only one sample that has log2CountArea asymmetric enough, I'll use sc method to determine the outliers in that metric

```{r}
sr_spe_list <- imap(sr_spe_list, \(spe, name){
  spe <- computeSpatialOutlier(spe, computeBy="Area_um", method="mc") # "mc" for asymmetric distribution of the choosen metric
  spe <- computeSpatialOutlier(spe, computeBy="log2CountArea", method="sc") # "sc" for symmetric distribution of the choosen metric
  spe
})
```

```{r}
outlier_summary_df <- imap_dfr(sr_spe_list, \(spe, name) {
  
  colData(spe) |>
    as.data.frame() |>
    select(Area_um_outlier_mc, log2CountArea_outlier_sc) |>
    
    # --- THE FIX IS HERE ---
    # Convert both columns to character to remove the special class
    mutate(across(everything(), as.character)) |>
    
    # Now pivot_longer will work correctly
    pivot_longer(
      cols = everything(),
      names_to = "metric",
      values_to = "outlier_type"
    ) |>
    
    count(metric, outlier_type, name = "count") |>
    
    mutate(sample = name)
})

outlier_summary_df |> 
  filter(
    metric == "Area_um_outlier_mc" & outlier_type %in% c("HIGH","LOW")
  ) |> 
  select(sample, count, outlier_type, metric)


outlier_summary_df |> 
  filter(
    metric == "log2CountArea_outlier_sc" & outlier_type == "LOW"
  ) |> 
  select(sample, count, outlier_type, metric)


```

Plots of the nuclei centroids colored by outliers info

```{r}

# "Area_um_outlier_mc"
# "log2CountArea_outlier_sc"


metric_to_plot = "log2CountArea_outlier_sc"
plots <- imap(sr_spe_list, \(spe, name){
  p <- plotCentroids(spe, colourBy = metric_to_plot, size = 0.5) +
          ggtitle(name)
  p
})
plots
# p1 <- wrap_plots(plots[c(1:12)], ncol = 4) + 
#   plot_annotation(title = paste0('Nuclei outliers - ', metric_to_plot))
# p2 <- wrap_plots(plots[c(13:24)], ncol = 4) + 
#   plot_annotation(title = paste0('Distribution of ',metric_to_plot,' Across All Samples. pt2'))
# 
# ggsave(paste0(pl_path,"area_outlier/", metric_to_plot, "_part1.png"), plot = p1, dpi = 300, units = "cm", width = 50, height = 30)
# ggsave(paste0(pl_path,"area_outlier/", metric_to_plot, "_part2.png"), plot = p2, dpi = 300, units = "cm", width = 50, height = 30)

iwalk(plots, function(plot, name){
 ggsave(paste0(pl_path,"CountArea_outlier/", name, ".png"), plot = plot, dpi = 300, units = "cm", width = 50, height = 30)
  #ggsave(paste0(pl_path,"area_outlier/", name, ".png"), plot = plot, dpi = 300, units = "cm", width = 50, height = 30)
})
```




