---
title: "Cell type annotation with RCTD"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(Banksy)
library(purrr)
library(DropletUtils)
library(dplyr)
library(ggplot2)
library(jsonlite)
library(patchwork)
library(scDblFinder)
library(SingleCellExperiment)
library(SpatialExperiment)
library(tidyr)
library(forcats)
library(spacexr)
library(SpotSweeper)
library(writexl)
library(ggspavis)
library(data.table)
```

# Deconvolution

## Data reading

Paths
```{r}
pl_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/RCTD_doublet/"
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = "/mnt/europa/valerio/repositories/cachetic_visiumHD/results/RCTD_doublet/"
```


SpatialExperiment of all merged samples
```{r}
load(paste0(Rd_path, "spe_list_nuclei.RData"))
```

Reference dataset from Elena
```{r}
sce_ref <- readRDS(ref_path)
```

## Robust Cell Type Decomposition (spacexr)

```{r rctd_function}
run_my_rctd <- function(spe, sce_ref, seed = 123, max_cores = 4, rctd_mode = "doublet", cell_type_col = "subclusters") {
  # --- 1. Setup and Deconvolution (as before) ---
  set.seed(seed)
  rownames(spe) <- make.unique(rownames(spe))
  
  # Prepare RCTD dataset
  rctd_data <- createRctd(
    spatial_experiment = spe,
    reference_experiment = sce_ref,
    cell_type_col = cell_type_col
  )
  
  # Run deconvolution
  results <- runRctd(rctd_data, rctd_mode = rctd_mode, max_cores = max_cores)
  
  # --- 2. Extract Requested Columns from colData ---
  cd <- colData(results)
  
  # Create the data.table by manually specifying each column.
  dt <- data.table(
    cell_id = rownames(cd),
    spot_class = cd$spot_class,
    first_type = cd$first_type,
    second_type = cd$second_type
  )
  # --- 3. Return the data.table ---
  return(dt)
}
```

Color vector and cell types
```{r}
# color_vector <- c(
#     "Endothelial" = "#0072B2",       # blu vivo
#     "FAPs" = "#009E73",              # verde vivo
#     "Immune_Cells" = "#D55E00",      # arancione-rosso
#     "MuSC" = "#E69F00",              # arancione
#     "Myonuclei_IIb" = "#CC79A7",     # magenta
#     "Myonuclei_IIx" = "#56B4E9",     # azzurro chiaro
#     "Myonuclei_IIx_IIa" = "#F0E442", # giallo
#     "Myonuclei_IIx_IIb" = "#9B59B6", # viola chiaro
#     "Myonuclei_MTJ" = "#0099B4",     # turchese
#     "Myonuclei_NMJ" = "#DDAA33",     # senape
#     "Myonuclei_Trim63" = "#CC3311",  # rosso scuro
#     "Nervous_System" = "#44AA99",    # verde acqua
#     "Pericyte" = "#AA4499",          # viola/rosa
#     "Smooth_Muscular" = "#332288",   # blu-viola scuro
#     "Tenocyte" = "#A3E635"           # verde lime
# )
# 
# cell_types <- c(
#   "Endothelial", "FAPs", "Immune_Cells", "MuSC",
#   "Myonuclei_IIb", "Myonuclei_IIx", "Myonuclei_IIx_IIa", "Myonuclei_IIx_IIb",
#   "Myonuclei_MTJ", "Myonuclei_NMJ", "Myonuclei_Trim63", "Nervous_System",
#   "Pericyte", "Smooth_Muscular", "Tenocyte"
# )
# save(color_vector, file = paste0(Rd_path, "color_cell_dec.RData"))
# save(cell_types, file = paste0(Rd_path, "cell_types.RData"))
```

Load color vector and cell types names vector
```{r}
load(paste0(Rd_path, "color_cell_dec.RData"))
load(paste0(Rd_path, "cell_types.RData"))
```

## All samples RCTD annotation

Run RCTD to all samples, the function return a dataframe with this three column:
 - spot_class: singlet, doublet, etc
 - first_type: of cell, primary cell type
 - second_type: second best guess
 
I removed blocco7 because after the filtering has too little genes in common with the reference

```{r}
spe_list_wb7 <- spe_list[!startsWith(names(spe_list), "blocco7")]
ann_list_RCTD <- imap(spe_list_wb7, \(spe, name){
  print(name)
  df <- run_my_rctd(spe, sce_ref, max_cores = 12, rctd_mode = "doublet", cell_type_col = "subclusters")
  df
  })

# salvo
# save(ann_list_RCTD, file = paste0(Rd_path, "annot_results_RCTD.RData"))
```

Add cell_type info to the spe_list_wb7
```{r}
ann_list_RCTD
spe_list_updated <- map2(spe_list_wb7, ann_list_RCTD, function(spe, ann_df) {
    # Find the matching order for cell IDs
    match_indices <- match(colnames(spe), ann_df$cell_id)
    # Add the 'first_type' column to colData, reordering based on the match
    spe$labels_rctd <- ann_df$first_type[match_indices]
    
    # Return the modified spe object
    return(spe)
  }
)
```


### Results

```{r}
lapply(ann_list_RCTD, function(elem) print(table(elem$spot_class)))
lapply(ann_list_RCTD, function(elem) print(table(elem$first_type)))
lapply(spe_list_updated, function(spe) {summary(as.factor(spe$first_type))})
```



```{r}
heatplot_doublet <- function(dt){
  # dt <- ann_list_RCTD$blocco1_c26STAT3
  
  plot_data <- dt |> 
    count(spot_class, first_type, name = "count") # We name the count column "count"
  
  # 2. Create the heatmap plot
  p <- ggplot(plot_data, aes(x = spot_class, y = first_type, fill = count)) +
    geom_tile(color = "white", lwd = 1) + # geom_tile creates the heatmap squares
    geom_text(aes(label = count), color = "black", size = 4) + # Add the count numbers on top
    scale_fill_gradient(low = "lightblue", high = "darkblue") + # Define the color scale
    labs(
      title = "Frequency of First Type within each Spot Class",
      x = "Spot Class",
      y = "First Type",
      fill = "Frequency"
    ) +
    theme_minimal() + # Use a clean theme
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 12),
      axis.text.y = element_text(size = 12),
      panel.grid = element_blank() # Remove grid lines for a cleaner look
    )
  return(p)
}
```

Heatmap of cell_type vs spot classification

```{r}
plots <- map(ann_list_RCTD, heatplot_doublet)

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path, name, "_spot_heatmap.png"), plot = plot, dpi = 700)
})
```



#### Plots

Load dataset with all annotation and redo the plots
```{r}
load(paste0(Rd_path, "spe_list_nuclei_ann.RData"))
```

Remove the blocco7
```{r}
spe_list_wb7 <- spe_list_up[!startsWith(names(spe_list_up), "blocco7_")]
```


Visualizzazione delle annotazioni sui tessuti, prima tutti i nuclei e relative
annotazioni e sotto solo i nuclei annotati, rimuovendo gli NA
```{r rctd_plot}

plots <- imap(spe_list_wb7, ~ {
  plotCoords(.x, annotate = "labels_rctd",
             point_size = 0.5,
             x_coord = "y", y_coord = "x") +
    scale_color_manual(values = color_vector) +
    theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    ) +
    ggtitle(.y)
})


iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path, "spat_annot/", name, "_annot.png"), plot = plot, dpi = 400, create.dir = TRUE)
})
```



```{r}
plots <- imap(spe_list_wb7, function(spe, nm) {
  # Determine where the annotation lives and subset accordingly
    keep <- !is.na(spe$labels_rctd)
    if (!any(keep)) return(NULL)
    p <- spe[, keep] |> 
      plotCoords(annotate = "labels_rctd",
                    point_size = 0.5, x_coord = "y", y_coord = "x") + 
      scale_color_manual(values = color_vector) +
      theme(legend.key.width  = unit(0.1, "lines"),
            legend.key.height = unit(1, "lines")
            ) + 
      ggtitle(nm)
})

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path, "spat_annot/", name, "_ann_noNA.png"), plot = plot, dpi = 700)
})
```


### Divided by cell type
```{r}
cell_types <- unique(spe_list_wb7$blocco1_c26STAT3$labels_rctd)[-1] # first is NA
plot_grids <- imap(spe_list_wb7, function(spe, spe_name) {

  # Inner loop to create a list of plots for each cell type
  list_of_plots <- map(cell_types, function(type) {
    
    # --- YOUR FIX IMPLEMENTED ---
    # 1. Create a local copy to be safe.
    spe_local <- spe
    
    # 2. Programmatically create a new column named after the content of `type`.
    # For example, if type is "Endothelial", this creates spe_local$Endothelial
    spe_local$is_here <- spe_local$labels_rctd == type
    keep <- !is.na(spe_local$labels_rctd)
    if (!any(keep)) return(NULL)
    
    p <- spe_local[, keep] |> plotCoords(x_coord = "y", y_coord = "x",
      annotate = "is_here", # This now correctly refers to the new column
      point_size = 0.05
    ) +
      scale_color_manual(values = c("TRUE" = "#CC3311", "FALSE" = "grey80"), name = "Present") +
      labs(title = type) # The subplot title is still the cell type name
    
    return(p)
  })
  
  # --- The rest of your code is correct ---
  
  # Combine the list of plots into a single grid
  final_grid <- wrap_plots(list_of_plots, nrow = 3, ncol = 5, guides = "collect") &
    theme(
      legend.key.width = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines"),
      plot.title = element_text(size = 8)
    )
  
  # Add a main title for the entire grid
  final_grid <- final_grid + plot_annotation(title = spe_name)
  
  return(final_grid)
})


iwalk(plot_grids, function(plot, name){
  # ggsave(paste0(pl_path,"spat_annot/by_type/", name, "_b1_miei.png"), plot = plot, dpi = 400, units = "cm", width = 50, height = 30)
  ggsave(paste0(pl_path,"spat_annot/by_type/", name, "_rctd_annot.png"), plot = plot, dpi = 400, units = "cm", width = 30, height = 20, create.dir = TRUE)
})
```

#### Barplot

```{r}
# Costruisci un data frame con tutte le proporzioni per ogni oggetto
prop_df_ordered <- imap_dfr(spe_list_wb7, function(spe, name) {
  tab <- base::table(spe$labels_rctd)
  prop <- prop.table(tab)
  tissue_type <- unique(spe$tissue_type)
  data.table(
    sample_id = name,
    cell_type = names(prop),
    proportion = as.numeric(prop),
    tissue_type = tissue_type
  )
}) |> 
  mutate(sample_id = fct_reorder(sample_id, tissue_type))


# orizzontale bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +
  coord_flip() +
  scale_x_discrete(name = "Sample") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +
  theme_minimal()

ggsave(paste0(pl_path, "cell_type_proportions_oriz.png"), width = 10, height = 6, dpi = 300)

# vertical bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +                             # geom_col uses the y value directly (identity)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +# use your named color_vector; otherwise use scale_fill_brewer()
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(paste0(pl_path, "cell_type_proportions_vert.png"), width = 10, height = 6, dpi = 300)

```

#### Barplot aggregato per condizione (SENZA BLOCCO 7)

```{r}
prop_df_condition <- imap_dfr(spe_list_updated, function(spe, name) {
  tibble(
    tissue_type = unique(spe$tissue_type),
    cell_type = spe$labels_rctd
  ) |> 
    filter(!is.na(cell_type))
}) |> 
  count(tissue_type, cell_type) |>
  group_by(tissue_type) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()

ggplot(prop_df_condition, aes(x = tissue_type, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per condizione",
    x = "Condizione (tissue_type)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(paste0(pl_path, "cell_type_proportions_by_condition.png"), width = 10, height = 6, dpi = 300)
```

## Tabelle proporzioni

```{r}
#per campione
celltype_table <- imap_dfr(spe_list_updated, function(spe, name) {
  as.data.frame(colData(spe)) |> 
    filter(!is.na(labels_rctd)) |>             # drop NA cell_type rows
    count(labels_rctd, name = "count")  |>      # tally per cell type
    mutate(proportion = count / sum(count),
           sample = name)                    # use imap name as sample id
})

celltype_matrix <- celltype_table |>
  select(sample, labels_rctd, proportion) |>
  pivot_wider(names_from = labels_rctd, values_from = proportion, values_fill = 0)

celltype_matrix

write_xlsx(celltype_matrix, paste0(res_path, "celltype_proportions_per_sample.xlsx"))
```


```{r}
#per tipo di campione (SENZA B7) ----------------------------------------------------
props_wide <- imap_dfr(spe_list_updated, function(spe, name) {
  tibble(
    tissue_type = rep(as.character(unique(spe$tissue_type)),
                      length.out = length(spe$labels_rctd)),
    cell_type   = as.character(spe$labels_rctd)
  )
}) |>
  count(tissue_type, cell_type, name = "n") |>
  group_by(tissue_type) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  select(tissue_type, cell_type, prop) |>
  pivot_wider(names_from = cell_type, values_from = prop, values_fill = 0)

write_xlsx(props_wide, path = paste0(res_path, "celltype_by_tissue_type.xlsx"))

```

# boxplot cell type proportions stratified by condition (SENZA BLOCCO7)
```{r}
prop_per_sample <- imap_dfr(spe_list_updated, function(spe, name) {
  cd <- as.data.frame(SummarizedExperiment::colData(spe))
  
  cd |> 
    filter(!is.na(labels_rctd)) |> 
    count(labels_rctd, name = "count") |> 
    mutate(proportion  = count / sum(count),
           sample      = name,
           tissue_type = paste(unique(as.character(cd$tissue_type)), collapse = ", "))
}) |> tibble()

ggplot(prop_per_sample, aes(x = tissue_type, y = proportion, fill = tissue_type)) +
  geom_boxplot(outlier.shape = 21) +
  facet_wrap(~ labels_rctd, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1b9e77", "c26" = "#d95f02")) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribuzione proporzioni cell_type per tissue_type",
       x = "Tissue type",
       y = "Proporzione")

ggsave(paste0(pl_path, "boxplot_celltype_prop_by_cond.png"), width = 10, height = 6, dpi = 300)
```

#Spatial relationship between trim and 2b myo
#questo in realtà bisognerebbe vederlo sulla segmentazione
```{r}

default_color <- "white"

your_color_vector <- setNames(
  rep(default_color, 16),
  c("Endothelial", "FAPs", "Immune_Cells", "MuSC", "Myonuclei_IIx",
    "Myonuclei_IIx_IIa", "Myonuclei_IIx_IIb", "Myonuclei_MTJ",
    "Myonuclei_NMJ", "Nervous_System", "Pericyte",
    "Smooth_Muscular", "Tenocyte",
    "Myonuclei_IIb", "Myonuclei_Trim63")
)

your_color_vector["Myonuclei_IIb"]     <- "blue"
your_color_vector["Myonuclei_Trim63"]  <- "red"
# 
# plot_list <- imap(spe_list_updated, function(spe, nm) {
#   plotCoords(spe, annotate = "labels_rctd", point_size = 0.5) +
#     scale_color_manual(values = your_color_vector, na.value = default_color) +
#     ggtitle(nm) +   # <- titolo con il nome del campione
#     theme(
#       legend.key.width  = unit(0.5, "lines"),
#       legend.key.height = unit(1, "lines"),
#       plot.title = element_text(hjust = 0.5)  # per centrare il titolo
#     )
# })


plots <- imap(spe_list_updated, function(spe, nm) {
  # Determine where the annotation lives and subset accordingly
    keep <- spe$labels_rctd %in% c("Myonuclei_IIb", "Myonuclei_Trim63")
    if (!any(keep)) return(NULL)
    p <- spe[, keep] |> 
      plotCoords(annotate = "labels_rctd", point_size = 0.5) +
      scale_color_manual(values = your_color_vector) +
      ggtitle(nm) +   # <- titolo con il nome del campione
      theme(
      legend.key.width  = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines"),
      plot.title = element_text(hjust = 0.5)  # per centrare il titolo
      )
})

iwalk(plots, function(plot, name){
  ggsave(paste0(pl_path,"spat_annot/Trim63vsIIb/", name, ".png"), plot = plot, dpi = 700)
})
```

# Final thing

Add the labels info to the spe_list with the SingleR ones

```{r}
load(paste0(Rd_path,"spe_list_nuclei_ann.RData"))
spe_list_updated <- map2(spe_list_up, ann_list_RCTD, function(spe, ann_df) {
    # Find the matching order for cell IDs
    match_indices <- match(colnames(spe), ann_df$cell_id)
    # Add the 'first_type' column to colData, reordering based on the match
    spe$labels_rctd <- ann_df$first_type[match_indices]
    # Return the modified spe object
    return(spe)
  }
)
```

