---
title: "PASTA exploration"
output: html_document
params:
  saves_path_stem: "PASTA/vario"
  labels_key: "labels_singler"
  spe_key: "sfe_list_annotated"
  #spe_key: "filtered_spe_annotated_28Nov"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(ggspavis)
library(ggnewscale)
library(patchwork)
library(SingleCellExperiment)
library(SpatialExperiment)
library(SpatialFeatureExperiment)
library(data.table)
library(purrr)
library(dplyr)
library(tidyr)
library(jsonlite)
library(SpaceTrooper)
library(sfarrow)
source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/pasta_fun.R")

# more from PASTA
library(spatstat)
library(rlang)
library(stats)
library(reshape2)
library(rgeoda)
library(Voyager)
library(scran)
library(tmap)
library(spdep)
library(stringr)
library(magrittr)
library(bluster)
library(dixon)
library(ExperimentHub)
library(scater)
library(spatstat.geom)
library(spatstat.explore)
library(spatialFDA)
library(sosta)
theme_set(theme_light())
```

# Plots post annotation

Paths
```{r}
pl_path = paste0("/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/", params$saves_path_stem,"/")
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = paste0("/mnt/europa/valerio/repositories/cachetic_visiumHD/results/", params$saves_path_stem, "/")
poly_path = "/mnt/europa/valerio/data/zarr_store/samples/"
```

Data
```{r}
color_vector <- readRDS(paste0(Rd_path, "color_cell_dec"))
sfe_list <- readRDS(paste0(Rd_path, params$spe_key))
#spe_list <- readRDS(paste0(Rd_path, "spe_list_nuclei_ann"))
# spe_list <- spe_list[stringr::str_detect(names(spe_list), "^blocco[12]")]
# saveRDS(filtered_spe, file = paste0(Rd_path, "filtered_spe_annotated_28Nov"))
# spe_list
```

Filtering bad tissue
```{r}
filtered_sfe <- sfe_list |> 
  map(function(sfe){
    sfe <- sfe[, !sfe$to_discard]
    if ("labels_rctd" %in% colnames(colData(sfe))) {
      cd <- colData(sfe)[, setdiff(colnames(colData(sfe)), "labels_rctd"), drop = FALSE]
      colData(sfe) <- cd
    }
    sfe
  })
filtered_sfe$blocco1_c26STAT3
```

```{r}
# filtered_sfe <- map(filtered_sfe, function(sfe) {
#   if ("labels_rctd" %in% colnames(colData(sfe))) {
#     cd <- colData(sfe)[, setdiff(colnames(colData(sfe)), "labels_rctd"), drop = FALSE]
#     colData(sfe) <- cd
#   }
#   sfe
# })

samples_sub <- list("blocco1_sham", "blocco2_c26", "blocco3_sham", "blocco4_c26")
filtered_sfe <- filtered_sfe[names(filtered_sfe) %in% as.vector(samples_sub)]
```


```{r}
sfe_combined <- do.call(cbind, filtered_sfe)
sfe_combined
```


Extract some samples and transform into ppp objects

```{r}
# define the samples that you want to compare
samples_sub <- list("blocco1_sham", "blocco2_c26", "blocco3_sham", "blocco4_c26")

# small helper function to extract the samples and convert them to `ppp` objects
select_samples <- function(sample, sfe) {
  sfe[, sfe$sample_id == sample] |>
    .ppp(marks = "labels_singler")
}
pp_ls <- map(samples_sub, select_samples, sfe_combined) |>
  setNames(samples_sub)

pp_ls
```

```{r}
pp_df_categorical <- lapply(samples_sub, function(x) {
  df <- pp_ls[[x]] |> as.data.frame()
  df$sample_id <- x
  return(df)
}) |> bind_rows()

```

Kest and Lest function for a single mark (MuSC)

```{r}
res_ls <- lapply(list('Kest', 'Lest'), function(fun) {
  res <- calcMetricPerFov(
    sfe_combined,
    'MuSC',
    subsetby = 'sample_id',
    fun = fun,
    marks = 'labels_singler',
    rSeq = NULL,
    by = c('sample_id'),
    ncores = 4
  )
  return(res)
})

p_ls <- lapply(res_ls, function(res) {
  plotMetricPerFov(
    res,
    theo = TRUE,
    correction = "iso",
    x = "r",
    imageId = 'sample_id',
    legend.position = "right"
  )
})
wrap_plots(p_ls, guides = 'collect')
```

# Local measurement

LiSA for MuSC

```{r}
pp <- subset(pp_ls[["blocco1_sham"]], marks %in% "MuSC")
L_musc_lisa <- localL(pp)

df <- as.data.frame(L_musc_lisa)
dfm <- reshape2::melt(df, "r")

get_sel <- dfm |>
  dplyr::filter(r > 0 & r < 50, variable != "theo") |>
  dplyr::mutate(sel = value) |>
  dplyr::select(variable, sel)

dfm <- dfm |> left_join(get_sel)

p <- ggplot(dfm, aes(
  x = r,
  y = value,
  group = variable,
  colour = sel
)) +
  geom_line(linewidth = 1) +
  scale_color_continuous(type = "viridis") +
  geom_vline(xintercept = 200) +
  theme(legend.position = "none") +
  ggtitle("LISA curves of sample blocco1_sham")
```

```{r}
p
```

Comparison between two cell types
```{r}
pp_df_trim_IIb <- pp_df_categorical[pp_df_categorical$marks %in% c("Myonuclei_Trim63", "Myonuclei_IIb"), ]

ggplot(pp_df_trim_IIb, aes(x, y, colour = marks)) +
  geom_point(size = 0.5) +
  facet_wrap( ~ sample_id, scales = "free") +
  theme(aspect.ratio = 1)
```

```{r}
res <- calcMetricPerFov(
  sfe_combined,
  c("Myonuclei_Trim63", "Myonuclei_IIb"),
  subsetby = 'sample_id',
  fun = 'Lcross',
  marks = 'labels_singler',
  rSeq = NULL,
  by = 'sample_id'
)


p <- plotMetricPerFov(
  res,
  theo = TRUE,
  correction = "iso",
  x = "r",
  imageId = 'sample_id',
  legend.position = "right"
)
```

```{r}
p
```


# Multivariate point processes

```{r}
setRiprasWindows <- function(pp){
  Window(pp) <- ripras(pp)
  return(pp)
}
#the entire point patterns with the ripras windows
pp <- lapply(pp_ls, setRiprasWindows)

separateMarks <- function(pp){
  #split the multitype point process into several single type processes
  ppls <- split(pp)
  return (ppls)
}
#the point patterns separated by their marks
pp_ls <- lapply(pp, separateMarks)
```

```{r}
library(sosta)
### code adapted from https://bioconductor.org/packages/release/bioc/vignettes/sosta/inst/doc/StructureReconstructionVignette.html ###

#subset to three images
spe_sub <- subset(sfe_combined, ,sample_id %in% samples_sub)

# first we select the intensity thresholds for the binarisation of the image
n <- estimateReconstructionParametersSPE(
    spe_sub,
    marks = "labels_singler",
    imageCol = "sample_id",
    markSelect = NULL,
    plotHist = TRUE
)

thresSPE <- mean(n$thres)
bndwSPE <- mean(n$bndw)

# calculate the observation window based on the density
allStructs <- reconstructShapeDensitySPE(
    spe_sub,
    marks = "labels_singler",
    imageCol = "sample_id",
    markSelect = NULL,
    bndw = bndwSPE,
    thres = thresSPE,
    nCores = 1
)

p_ls <- lapply(samples_sub, function(elem){
    spe_sub <- subset(sfe_combined,,sample_id == elem)
    allStructs_sub <- subset(allStructs, sample_id == elem)
    p <- cbind(colData(spe_sub), spatialCoords(spe_sub)) |>
    as.data.frame() |>
    ggplot(aes(x = y, y = x)) +
    geom_point(size = 0.25) +
    geom_sf(
      data = allStructs_sub,
      fill = NA,
      color = "red",
      inherit.aes = FALSE, # this is important
      linewidth = 0.8
    ) +
    facet_grid(~sample_id)+
    theme_light()
    return(p)
})
patchwork::wrap_plots(p_ls)
```

Intensity estimate

```{r}
ppDens <- SPE2ppp(sfe_combined,
marks = "labels_singler", imageCol = "sample_id",
imageId = "blocco1_sham"
)
#dens_im <- density()
dens_im <- density(unmark(ppDens))

# calculate the scale as we have less points compared to unmarked
scale <- sum(dens_im$v) / (dens_im$xstep * dens_im$ystep)

# scale the density
dens_im_scaled <- dens_im * scale

plot(dens_im_scaled)
```

# Confronto tra trim63 e IIb nei c26

```{r}
filtered_sfe <- sfe_list[endsWith(names(sfe_list),"c26")]
names(filtered_sfe)

filtered_sfe <- filtered_sfe |> 
  map(function(sfe){
    sfe <- sfe[, !sfe$to_discard]
    if ("labels_rctd" %in% colnames(colData(sfe))) {
      cd <- colData(sfe)[, setdiff(colnames(colData(sfe)), "labels_rctd"), drop = FALSE]
      colData(sfe) <- cd
    }
    spatialCoordsNames(sfe) <- c("x", "y")
    sfe
  })
```

```{r}
sfe_combined <- do.call(cbind, filtered_sfe)
sfe_combined

```

Window definition

```{r}
n <- estimateReconstructionParametersSPE(
    sfe_combined,
    marks = "labels_singler",
    imageCol = "sample_id",
    markSelect = NULL,
    plotHist = TRUE
)

thresSPE <- mean(n$thres)
bndwSPE <- mean(n$bndw)

# calculate the observation window based on the density
allStructs <- reconstructShapeDensitySPE(
    sfe_combined,
    marks = "labels_singler",
    imageCol = "sample_id",
    markSelect = NULL,
    bndw = bndwSPE,
    thres = thresSPE,
    nCores = 4
)

p_ls <- lapply(names(filtered_sfe), function(elem){
    sfe_sub <- subset(sfe_combined,,sample_id == elem)
    allStructs_sub <- subset(allStructs, sample_id == elem)
    p <- cbind(colData(sfe_sub), spatialCoords(sfe_sub)) |>
    as.data.frame() |>
    ggplot(aes(x = x, y = y)) +
    geom_point(size = 0.25) +
    geom_sf(
      data = allStructs_sub,
      fill = NA,
      color = "red",
      inherit.aes = FALSE, # this is important
      linewidth = 0.8
    ) +
    facet_grid(~sample_id)+
    theme_light()
    return(p)
})
patchwork::wrap_plots(p_ls)
```

# estimate the intensity 

```{r}

ppDens_list <- map(sfe_combined$sample_id, function(sample){
  ppDens <- SPE2ppp(sfe_combined,
      marks = "labels_singler", imageCol = "sample_id", imageId = sample
  )
  dens_im <- density(unmark(ppDens))
  scale <- sum(dens_im$v) / (dens_im$xstep * dens_im$ystep)
  dens_im_scaled <- dens_im * scale
  dens_im_scaled
})
# 
# ppDens <- SPE2ppp(sfe_combined,
#   marks = "labels_singler", imageCol = "sample_id", imageId = "blocco2_c26"
# )
# #dens_im <- density()
# dens_im <- density(unmark(ppDens))
# 
# # calculate the scale as we have less points compared to unmarked
# scale <- sum(dens_im$v) / (dens_im$xstep * dens_im$ystep)
# 
# # scale the density
# dens_im_scaled <- dens_im * scale
# plot(dens_im_scaled)
```

```{r}
samples_sub <- names(filtered_sfe)
# small helper function to extract the samples and convert them to `ppp` objects
select_samples <- function(sample, sfe) {
  sfe[, sfe$sample_id == sample] |>
    .ppp(marks = "labels_singler")
}
pp_ls <- map(samples_sub, select_samples, sfe_combined) |>
  setNames(samples_sub)

pp_ls
```

Add the new windows

```{r}
# Update windows using purrr
pp_ls_windowed <- imap(pp_ls, function(pp, name) {
  
  # 1. Filter polygons for this sample ID
  sample_polys <- allStructs[allStructs$sample_id == name, ]
  
  # 2. If polygons exist, update the window; otherwise return original pp
  if (nrow(sample_polys) > 0) {
    # Convert sf to owin (automatically unions multiple polygons)
    new_window <- as.owin(sample_polys)
    
    # Apply new window (subsets points inside the polygon)
    return(pp[new_window])
  } else {
    warning(paste("No polygons found for:", name))
    return(pp)
  }
})

# Verify
print(pp_ls_windowed$blocco2_c26)
plot(pp_ls_windowed$blocco2_c26)
```


Kernel density estimation - non parametric intensity estimation

```{r}
dens_trim63_list <- pp_ls_windowed |> 
  map(function(pp){
    pp <- pp[pp$marks=="Myonuclei_Trim63"]
    bwidth <- bw.diggle(pp)
    dens <- density.ppp(pp, sigma = bwidth)
    dens
  })
iwalk(dens_trim63_list, function(dens, name){
  plot(dens, main = paste0("Kernel Density of Myonuclei Trim63 - sample ", name))
})
```

```{r}
dens_IIb_list <- pp_ls_windowed |> 
  map(function(pp){
    pp <- pp[pp$marks=="Myonuclei_IIb"]
    bwidth <- bw.diggle(pp)
    dens <- density.ppp(pp, sigma = bwidth)
    dens
  })
iwalk(dens_IIb_list, function(dens, name){
  plot(dens, main = paste0("Kernel Density of Myonuclei IIb - sample ", name))
})
```

```{r}
intensityPointProcess <- function(pp,mark) if(mark) intensity(pp) else sum(intensity(pp))

map(pp_ls_windowed, function(pp) {intensityPointProcess(pp, mark = TRUE) |> round(6)})
```


## Quadrat counting

```{r}
Q5 <- quadratcount(pp_ls$blocco2_c26, nx=8, ny=8)
plot(unmark(pp_ls$blocco2_c26), main='Unmarked Point Pattern Quadrats')
plot(Q5, col='red', add=TRUE)
```

```{r}
val <- quadrat.test(subset(pp_ls$blocco2_c26, marks == 'Myonuclei_IIb'), 8, alternative="regular", method="MonteCarlo")
val
```

