---
title: "Esplorazione Bin"
author: "Valerio Reffo"
date: "2026-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(ggplot2)
library(ggspavis)
library(jsonlite)
library(anndataR)
#library(scuttle)
library(SpatialExperiment)
library(data.table)
# to read spatialdata objects
library(ggnewscale)
#library(SpatialData)
library(SingleCellExperiment)
# library(SingleR)
library(BiocParallel)
#source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")

library(dplyr)
library(tidyr)
library(forcats) # For ordering plots

```

Leggiamo i dati di emma sui 16 bin (e forse anche 8 bin se li ho) e vediamo un po' di statistiche su questi markers.

```{r}
spe_list <- readRDS("~/data/Rdata/emma_16um_bin_annotated_spelist.RDS")
spe <-spe_list$sham_b1

marker_genes <- readRDS("~/data/Rdata/marker_genes_list_singleR_annot")
marker_genes

color_vector <- readRDS("~/data/Rdata/color_cell_dec")

```

```{r}
spe
```

# Analysis of RCTD results

There are a lot of Endothelial, but why? what is the distribution of the probability for each bin to be a Endothetial? 

```{r}
cd_df <- colData(spe) |> as_tibble()

```

```{r}
cd_df |> 
  filter(cell_type == "Endothelial") |> 
  summarise(mean = mean(Endothelial),
            sd = sd(Endothelial),
            min = min(Endothelial),
            max = max(Endothelial))

cd_df |> 
  filter(cell_type == "Endothelial", Endothelial == 0) |> 
  summarise(n = n())
cd_df |> 
  filter(cell_type == "Endothelial", Endothelial == 1) |> 
  summarise(n = n())
cd_df |> 
  filter(
    cell_type == "Endothelial", 
    between(Endothelial, 0.5, 0.99)
  ) |> 
  summarise(n = n())
cd_df |> 
  filter(
    cell_type == "Endothelial", 
    between(Endothelial, 0.0001, 0.5)
  ) |> 
  summarise(n = n())
```

```{r}
# 1. Filter only the bins assigned as Endothelial
p_hist <- as.data.frame(cd_df) |> 
  filter(cell_type == "Endothelial") |>
  ggplot(aes(x = Endothelial)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Distribution of Endothelial Scores",
    subtitle = "Only including bins classified as 'Endothelial'",
    x = "Endothelial Score (0-1)",
    y = "Number of Bins"
  )

print(p_hist)
```



Second highest scores among Endothelial defined cell types
```{r}
# Define all your cell type score columns
score_cols <- c("Endothelial", "FAPs", "Immune_Cells", "MuSC", 
                "Myonuclei_IIb", "Myonuclei_IIx", "Myonuclei_IIx_IIa", 
                "Myonuclei_IIx_IIb", "Myonuclei_MTJ", "Myonuclei_NMJ", 
                "Myonuclei_Trim63", "Nervous_System", "Pericyte", 
                "Smooth_Muscular", "Tenocyte")

# Extract the second highest score for each Endothelial bin
second_highest_df <- cd_df |> 
  filter(cell_type == "Endothelial", Endothelial != 0) |>
  select(barcode, all_of(score_cols)) |> 
  pivot_longer(cols = -barcode, names_to = "candidate_type", values_to = "score") |>
  # Group by each individual bin
  group_by(barcode) |>
  # Sort scores from highest to lowest inside each bin
  arrange(desc(score)) |>
  # Grab the 2nd row (which is the second highest score)
  slice(2) |> 
  ungroup()

# Plot: WHICH cell types are most frequently the "second choice"?
p_2nd_identity <- ggplot(second_highest_df, aes(x = fct_infreq(candidate_type))) +
  geom_bar(fill = "darkorange", color = "black", alpha = 0.8) +
  theme_minimal() +
  coord_flip() + # Flip for readable names
  labs(
    title = "What is sharing space with Endothelial cells?",
    subtitle = "Most common 'second-highest' cell types in Endothelial bins",
    x = "Cell Type",
    y = "Count of Bins"
  )

print(p_2nd_identity)

```

```{r}
plot_confidence <- cd_df |> 
  filter(cell_type == "Endothelial", Endothelial != 0) |>
  left_join(second_highest_df |> select(barcode, second_type = candidate_type, second_score = score), by = "barcode") |>
  mutate(Confidence_Delta = Endothelial - second_score) |> 
  ggplot(aes(x = Confidence_Delta)) +
  geom_histogram(bins = 30, fill = "purple", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Assignment Confidence (Endothelial Score - 2nd Highest Score)",
       x = "Delta (Higher = Purer Bin, Lower = Mixed Bin)")
plot_confidence
```

```{r}
weird_endo <- cd_df |> 
  filter(cell_type == "Endothelial", Endothelial == 0)

# 2. Check the MAXIMUM score for all other cell types in these bins
cat("\n--- MAXIMUM SCORE IN OTHER CELL TYPES ---\n")
weird_endo |> 
  select(FAPs:Tenocyte) |> 
  summarise(across(everything(), max)) |> 
  print(width = Inf)

# 3. Check HOW MANY of these 5700 bins have a score > 0 for the other cell types
cat("\n--- NUMBER OF BINS WITH SCORE > 0 ---\n")
weird_endo |> 
  select(FAPs:Tenocyte) |> 
  summarise(across(everything(), ~ sum(. > 0))) |> 
  print(width = Inf)
```

```{r}
# 2. Reshape and calculate the summary
zero_score_summary <- cd_df |> 
  # Keep only the essential columns (barcode, assigned cell_type, and the score columns)
  select(barcode, assigned_type = cell_type, all_of(score_cols)) |> 
  
  # Pivot the score columns into two new columns: 'score_type' and 'score_value'
  pivot_longer(
    cols = all_of(score_cols), 
    names_to = "score_type", 
    values_to = "score_value"
  ) |> 
  
  # KEEP ONLY the rows where the assigned type matches the score column name
  filter(assigned_type == score_type) |> 
  
  # Group by the assigned cell type to calculate our stats
  group_by(assigned_type) |> 
  
  # Calculate how many total bins, and how many have a score of exactly 0
  summarise(
    total_bins = n(),
    bins_with_zero_score = sum(score_value == 0, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  
  # Add the percentage for context
  mutate(percentage_zero = (bins_with_zero_score / total_bins) * 100) |> 
  
  # Sort so the worst offenders are at the top
  arrange(desc(bins_with_zero_score))

# View the result
print(zero_score_summary)
```
Looks like is a unique problem of the Endothelial and MuSC (which are manually annotated i guess)

```{r}
cd_df <- cbind(cd_df, spatialCoords(spe))
valid_barcodes <- cd_df |> 
  select(barcode, cell_type, all_of(score_cols)) |> 
  pivot_longer(
    cols = all_of(score_cols), 
    names_to = "score_type", 
    values_to = "score_value"
  ) |> 
  # Keep only rows where the assignment matches the score column, AND the score is > 0
  filter(cell_type == score_type, score_value > 0) |> 
  pull(barcode) # This pulls just the barcodes into a simple vector

# 3. Filter your main dataframe using %in%
cd_df_clean <- cd_df |> 
  filter(barcode %in% valid_barcodes)

# 4. Plot the cleaned spatial map!
p_clean_map <- ggplot(cd_df_clean, aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = cell_type)) +
  geom_point(size = 0.6) +
  scale_y_reverse() + # Keep the tissue right-side up
  scale_color_manual(values = color_vector) +
  theme_void() +
  coord_fixed() +
  labs(
    title = "Spatial Map of Confident Cell Types",
    subtitle = paste0("Excluded ", nrow(cd_df) - nrow(cd_df_clean), " bins with an assigned score of 0"),
    color = "Cell Type"
  ) +
  # This makes the dots in the legend bigger so they are easier to read
  guides(color = guide_legend(override.aes = list(size = 4)))

print(p_clean_map)
```


```{r}
cell_proportions <- cd_df_clean |> 
  count(cell_type) |> 
  mutate(
    proportion = n / sum(n),
    percentage = proportion * 100
  ) |> 
  # Sort so the most abundant cell type is at the top of the plot
  arrange(desc(percentage))

# plot
p_stacked <- cd_df_clean |> 
  ggplot(aes(x = "Filtered Tissue", fill = cell_type)) +
  
  # Removed color = "white" so the bars touch seamlessly
  geom_bar(position = "fill", width = 0.4) +
  
  # Applies your custom color palette
  scale_fill_manual(values = color_vector) +
  
  theme_minimal() +
  labs(
    title = "Composition of Confident Cell Types",
    subtitle = paste0("Total valid bins: ", nrow(cd_df_clean)),
    x = NULL, 
    y = "Proportion (0 to 1)",
    fill = "Cell Type"
  ) +
  theme(
    axis.text.x = element_text(size = 14, face = "bold"), 
    panel.grid.major.x = element_blank(), 
    panel.grid.minor.x = element_blank()
  )

print(p_stacked)
```

```{r}
cd_df_analyzed <- cd_df |> 
  rowwise() |> 
  mutate(
    # What was the score of the cell_type it was originally assigned?
    original_score = c_across(all_of(score_cols))[match(cell_type, score_cols)],
    
    # What is the highest score across ALL cell types for this bin?
    max_score = max(c_across(all_of(score_cols))),
    
    # WHICH cell type has that highest score?
    best_type = score_cols[which.max(c_across(all_of(score_cols)))]
  ) |> 
  ungroup()

# 3. Print the Diagnostic Summary!
cat("\n--- WHAT IS INSIDE THE SCORE=0 BINS? ---\n")
cd_df_analyzed |> 
  filter(original_score == 0) |> 
  mutate(
    # Group them into "Completely Empty" vs "Mislabeled"
    status = if_else(max_score == 0, 
                     "All 15 scores are 0", 
                     paste("Actually belongs to:", best_type))
  ) |> 
  count(assigned_as = cell_type, status, name = "number_of_bins") |> 
  arrange(desc(number_of_bins)) |> 
  print(n = 50) # Prints the top 50 rows so you can see where they went
```

Create a new cell type column

```{r}
# 2. Define the function
fix_spe_celltypes <- function(spe, score_columns, cell_type_col = "cell_type") {
  
  # Extract colData as a dataframe
  cd <- as.data.frame(colData(spe))
  
  # Convert just the score columns into a numeric matrix (for extreme speed)
  score_mat <- as.matrix(cd[, score_columns])
  
  # A. Find the MAXIMUM score for each bin
  max_score <- apply(score_mat, 1, max)
  
  # B. Find WHICH cell type has that maximum score using max.col()
  best_type <- score_columns[max.col(score_mat, ties.method = "first")]
  
  # C. Find the score of the CURRENTLY assigned cell type
  # We find the column index of the assigned cell type, then extract the exact score
  orig_col_idx <- match(as.character(cd[[cell_type_col]]), score_columns)
  orig_score <- score_mat[cbind(1:nrow(score_mat), orig_col_idx)]
  
  # ---------------------------------------------------------
  # Apply your logic to create new_cell_type
  # ---------------------------------------------------------
  new_cell_type <- as.character(cd[[cell_type_col]])
  
  # Logic 1: If original score is 0, but another type is > 0 -> Reassign to best_type
  reassign_mask <- which(orig_score == 0 & max_score > 0)
  new_cell_type[reassign_mask] <- best_type[reassign_mask]
  
  # Logic 2: If EVERY score is 0 -> Assign NA
  na_mask <- which(max_score == 0)
  new_cell_type[na_mask] <- NA
  
  # Assign the new column back directly into the SPE object's colData
  spe$new_cell_type <- new_cell_type
  
  return(spe)
}
```

```{r}
spe <- fix_spe_celltypes(spe, score_cols)
```


# Analisi per tutta la lista dei campioni

```{r}
zeroscore_fun <- function(spe){
  cd_df <- colData(spe) |> as_tibble()
  
  # zero score summary info
  zero_score_summary <- cd_df |> 
    # Keep only the essential columns (barcode, assigned cell_type, and the score columns)
    select(barcode, assigned_type = cell_type, all_of(score_cols)) |> 
    
    # Pivot the score columns into two new columns: 'score_type' and 'score_value'
    pivot_longer(
      cols = all_of(score_cols), 
      names_to = "score_type", 
      values_to = "score_value"
    ) |> 
    
    # KEEP ONLY the rows where the assigned type matches the score column name
    filter(assigned_type == score_type) |> 
    
    # Group by the assigned cell type to calculate our stats
    group_by(assigned_type) |> 
    
    # Calculate how many total bins, and how many have a score of exactly 0
    summarise(
      total_bins = n(),
      bins_with_zero_score = sum(score_value == 0, na.rm = TRUE),
      .groups = "drop"
    ) |> 
    
    # Add the percentage for context
    mutate(percentage_zero = (bins_with_zero_score / total_bins) * 100) |> 
    
    # Sort so the worst offenders are at the top
    arrange(desc(bins_with_zero_score))
  
  # new cell types column to add to colData

    
  
  return(list(zeroScoreSummary = zero_score_summary, add2coldata = new_cell_types))
}

zeroscore_list <- map(spe_list, zeroscore_fun)
walk(zeroscore_list, \(el) {print(el$percentage_zero[1])})
```

Add new cell type column to each spe sample
```{r}
spe_list <- map(spe_list, \(spe) {spe <- fix_spe_celltypes(spe, score_cols)})

map(spe_list, \(x){table(x$new_cell_type)})
```

# plots of the differences

```{r}
prop_df_ordered <- imap_dfr(spe_list, function(spe, name) {
  tab <- base::table(colData(spe)[["new_cell_type"]])
  prop <- prop.table(tab)
  tissue_type <- unique(spe$tissue_type)
  data.table(
    sample_id = name,
    cell_type = names(prop),
    proportion = as.numeric(prop),
    tissue_type = as.character(tissue_type)
  )
}) |> 
  # lower case otherwise S is before f and m
  arrange(stringr::str_to_lower(tissue_type), sample_id) |>
  mutate(sample_id = fct_inorder(sample_id))

# vertical bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +                             # geom_col uses the y value directly (identity)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +# use your named color_vector; otherwise use scale_fill_brewer()
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave("~/repositories/cachetic_visiumHD/figures/output_R/emma_RCTD_by_me/cell_type_proportions_vert.png", width = 10, height = 6, dpi = 300)
```

By conditions
```{r}
prop_df_condition <- imap_dfr(spe_list, function(spe, name) {
  data.table(
    tissue_type = as.character(spe$tissue_type),
    cell_type = colData(spe)[["new_cell_type"]]
  ) |> 
    filter(!is.na(cell_type))
}) |> 
  arrange(stringr::str_to_lower(tissue_type)) |> 
  dplyr::count(tissue_type, cell_type) |>
  group_by(tissue_type) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()


ggplot(prop_df_condition, aes(x = tissue_type, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per condizione",
    x = "Condizione (tissue_type)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("~/repositories/cachetic_visiumHD/figures/output_R/emma_RCTD_by_me/cell_type_proportions_by_condition.png", width = 10, height = 6, dpi = 300)
```

## Cell type proportions stratified by condition

```{r boxplot-chunk, eval = !params$partial_data}
# This chunk will not run if partial_data == FALSE
prop_per_sample <- imap_dfr(spe_list, function(spe, name) {
  cd <- as.data.table(SummarizedExperiment::colData(spe))
  cd |>
    filter(!is.na(.data[["new_cell_type"]])) |>
    dplyr::count(.data[["new_cell_type"]], name = "count") |>
    mutate(proportion  = count / sum(count),
           sample      = name,
           tissue_type = paste(unique(as.character(cd$tissue_type)), collapse = ", ")) |> 
    dplyr::rename(cell_type = all_of("new_cell_type"))
})

ggplot(prop_per_sample, aes(x = tissue_type, y = proportion, fill = tissue_type)) +
  geom_boxplot(outlier.shape = 21) +
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1b9e77", "c26" = "#d95f02")) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribuzione proporzioni cell_type per tissue_type",
       x = "Tissue type",
       y = "Proporzione")

ggsave("~/repositories/cachetic_visiumHD/figures/output_R/emma_RCTD_by_me/boxplot_celltype_prop_by_cond.png", width = 10, height = 6, dpi = 300)
```




#------------------------------------------------------------------------------#
# Esploration of some genes

```{r}
marker_genes
genes_of_interest <- c("Pax7", "Col6a3")
```

```{r}
coords_df <- as.data.frame(spatialCoords(spe))
  
# 3. Extract the counts for just these genes, and log-transform them
# (If you already have a "logcounts" assay, use assay(spe, "logcounts") instead of log1p)
exprs_matrix <- as.matrix(assay(spe, "logcounts")[genes_of_interest, , drop = FALSE])
exprs_df <- as.data.frame(t(exprs_matrix)) # Log transform for better plotting

# 4. Extract cell metadata (assuming your cell types are in a column named 'Cell_Type')
# Change "Cell_Type" to whatever your column is actually named in colData(spe)!
meta_df <- as.data.frame(colData(spe))

# 5. Bind it all together
plot_df <- cbind(coords_df, exprs_df, meta_df)

# 6. Make a 'long' version for facet_wrap plotting
plot_df_long <- plot_df |> 
  pivot_longer(cols = all_of(genes_of_interest), 
               names_to = "Gene", 
               values_to = "Log_Expression")
```

Distribution of Endothelial markers
```{r}
p1 <- plot_df_long |> 
  # filter(sample_id == "YOUR_SAMPLE_NAME") |> # Uncomment this if you have multiple tissues!
  # filter(Log_Expression > 0) |> # Uncomment this to hide cells with 0 expression
  ggplot(aes(x = pxl_col_in_fullres, y = pxl_row_in_fullres, color = Log_Expression)) +
  geom_point(size = 0.5) +
  scale_y_reverse() + # Keeps the tissue right-side up!
  scale_color_viridis_c(option = "magma") +
  facet_wrap(~Gene) +
  theme_void() + 
  coord_fixed() + 
  labs(title = "Spatial Distribution of MuSC Markers", color = "logcounts")

print(p1)
```

Violin plots
```{r}
p2 <- ggplot(plot_df_long, aes(x = cell_type, y = Log_Expression, fill = cell_type)) +
  geom_violin(scale = "width", alpha = 0.7, color = NA) +
  # Adds tiny dots to show the actual cells inside the violins
  geom_jitter(size = 0.5, alpha = 0.2, width = 0.2) +
  facet_wrap(~Gene, ncol = 1, scales = "free_y") + # ncol=1 stacks them vertically for easier reading
  theme_minimal() +
  labs(title = "Expression by Cell Type", x = NULL, y = "Log(Counts + 1)") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
    legend.position = "none" # Hide the legend to save space
  )

print(p2)

# saving
file_path <- paste0("/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/emma_RCTD_by_me/violin_plots/MuSC_manual_celltypes.png")
ggsave(
      filename = file_path, 
      plot = p2, 
      width = 10, 
      height = 7, 
      dpi = 300, # High resolution for publications
      bg = "white" # Prevents transparent backgrounds
)
```

Function to do violin plot for everybody
```{r}
plot_marker_violins <- function(spe, marker_list, cell_type_col = "new_cell_type") {
  
  # Ensure the cell_type column exists and handle potential NAs
  if (!cell_type_col %in% colnames(colData(spe))) {
    stop(paste("Column", cell_type_col, "not found in colData(spe)."))
  }
  
  # Store the plots in a list so you can view/save them later
  plot_list <- list()
  
  # Loop through each cell type in your marker list
  for (cell_type_name in names(marker_list)) {
    
    # 1. Get the requested genes for this cell type
    genes_of_interest <- marker_list[[cell_type_name]]
    
    # Check if genes actually exist in the SPE object to prevent crashing
    valid_genes <- intersect(genes_of_interest, rownames(spe))
    
    if (length(valid_genes) == 0) {
      warning(paste("None of the genes for", cell_type_name, "were found in the SPE object. Skipping."))
      next # Skip to the next cell type
    }
    
    # 2. Extract Logcounts
    exprs_matrix <- as.matrix(assay(spe, "logcounts")[valid_genes, , drop = FALSE])
    exprs_df <- as.data.frame(t(exprs_matrix))
    
    # 3. Extract Metadata (we only need the cell type column to keep it fast)
    meta_df <- as.data.frame(colData(spe)[, cell_type_col, drop = FALSE])
    
    # 4. Bind and reshape
    plot_df <- cbind(meta_df, exprs_df) |> 
      # Drop NAs in cell type so they don't plot as a weird "NA" column
      filter(!is.na(.data[[cell_type_col]])) |> 
      pivot_longer(
        cols = all_of(valid_genes), 
        names_to = "Gene", 
        values_to = "Log_Expression"
      )
    
    # 5. Create the plot!
    p <- ggplot(plot_df, aes(x = .data[[cell_type_col]], y = Log_Expression, fill = .data[[cell_type_col]])) +
      geom_violin(scale = "width", alpha = 0.7, color = NA) +
      # Tiny dots to show density. (Rasterizing or alpha helps if you have 100k+ cells)
      geom_jitter(size = 0.1, alpha = 0.1, width = 0.2, color = "black") +
      facet_wrap(~Gene, ncol = 1, scales = "free_y") + 
      theme_minimal() +
      labs(
        title = paste("Marker Genes for:", cell_type_name), 
        x = NULL, 
        y = "Log Expression"
      ) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
        legend.position = "none",
        plot.title = element_text(face = "bold", size = 14)
      )
    
    # Save the plot into the list, using the cell type as the name
    plot_list[[cell_type_name]] <- p
  }
  
  return(plot_list)
}
```


```{r}
# This runs the loop and saves all the plots into 'all_violins'
all_violins <- plot_marker_violins(spe, marker_genes, cell_type_col = "new_cell_type")

print(all_violins$Endothelial)

walk(all_violins, print)
```

MuSC markers with old cell type annotations
```{r}
all_violins <- plot_marker_violins(spe, marker_genes, cell_type_col = "new_cell_type")
```


Save the plotting

```{r}
# This will run the loop and save everything automatically
imap(all_violins, \(one_plot, one_name) {

  markers_one_type <- marker_genes[[one_name]]
  dynamic_height <- 3 + (length(markers_one_type) * 2)
  file_path <- paste0("/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/emma_RCTD_by_me/violin_plots/", one_name, ".png")
  ggsave(
      filename = file_path, 
      plot = one_plot, 
      width = 10, 
      height = dynamic_height, 
      dpi = 300, # High resolution for publications
      bg = "white" # Prevents transparent backgrounds
    )
  
  })
```

```{r}
scater::plotDots(
  spe,
  features = marker_genes,
  group = "clusters",
  assay.type = "logcounts"
)
```


# -----------------------------------------------------------------------------#

Esploration of the dataset

```{r}
spe
```

```{r}
location <- rowRanges(spe)
is.mito <- any(seqnames(location)%in% c("MT", "Mt", "mt"))

sce.416b <- scuttle::addPerCellQCMetrics(spe, subsets=list(Mito=is.mito))
colnames(colData(spe))
```

