---
title: "Data reading"
author: "Valerio Reffo"
date: "2025-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(jsonlite)
library(anndataR)
library(SummarizedExperiment)
library(scuttle)
# to read spatialdata objects
library(ggnewscale)
library(SpatialData)
library(SingleCellExperiment)
```

# Concatenated samples

Al momento non è disponibile un file concatenato perché non riesco a farlo funzionare in python. Si può fare da R molto più 
tranquillamente.
```{r concat_data}
#(sdata <- readSpatialData("/mnt/europa/valerio/data/zarr_store/concat_all_samples_almost.zarr", anndataR=TRUE))

```

# List of all samples

This dictionary has all the info needed to read the data
```{r samples_dict, include=FALSE}
samples_dict <- fromJSON("/mnt/europa/valerio/repositories/cachetic_visiumHD/json/blocco_sample_bbox_dict.json")
```

Vector with samples name
```{r }
samples_key <- lapply(samples_dict, function(blocco) {
  lapply(blocco, function(sample) {
    return(sample$sample_key)
  })
}) |> unlist(use.names = FALSE)
samples_key
```

```{r reading_function, include=FALSE}
read_samples_list <- function(samples)
{
  spe_list <- lapply(samples, function(sample) {
    sdata <- readSpatialData(paste0("/mnt/europa/valerio/data/zarr_store/samples/", sample, ".zarr"), anndataR=TRUE)
    tables(sdata)$nuclei_counts
  })
  names(spe_list) = samples
  return(spe_list)
}
```

```{r samples_list}
spe_list = read_samples_list(samples_key)
spe_list
```


#-----------------------------------------------------------------------------

# Manual blasfemy
Nothing is working so I'll read data as single samples SingleCellExperiment
load spatial coordinates as csv and then build the objects I need manually.

(now is working - i need not only spatialdata to read spatialdata but also sce library)

```{r}
read_samples_adatas <- function(samples)
{
  sce_list <- lapply(samples, function(sample) {
    h5ad_path <- paste0("/mnt/europa/valerio/data/zarr_store/adatas/", sample, ".h5ad")
    sce <- read_h5ad(h5ad_path, to = "SingleCellExperiment")
  })
  names(sce_list) = samples
  return(sce_list)
}
```

```{r}
sce_list = read_samples_adatas(samples_key)
#sce_list
```

# Preparation of the different objects

## Single Cell Experiment
Adding some useful information to the colData and filtering low genes/counts nuclei

```{r}

for(nm in names(sce_list)) {
  blocco <- strsplit(nm, "_")[[1]][1]
  sampletype <- strsplit(nm, "_")[[1]][2]
  colData(sce_list[[nm]])$blocco <- blocco
  colData(sce_list[[nm]])$sampletype <- sampletype
  colData(sce_list[[nm]])$ncounts <- colSums(assay(sce_list[[nm]]))
  colData(sce_list[[nm]])$ngenes <- Matrix::colSums(assay(sce_list[[nm]]) > 0)
  sce_list[[nm]] <- sce_list[[nm]][, colData(sce_list[[nm]])$ncounts >= quantile(colData(sce_list[[nm]])$ncounts, 0.05) &
                       colData(sce_list[[nm]])$ngenes >= quantile(colData(sce_list[[nm]])$ngenes, 0.05)]
}
```

## Single sce object
```{r}
sce_merged <- do.call(cbind, sce_list)
sce_merged
```

Ok le dimensioni corrispondono

```{r}
sum(sapply(sce_list, ncol))
dim(sce_merged)
```


## Pseudobulk

```{r}
pb_list <- lapply(sce_list, function(sce) {
  aggregateAcrossCells(sce, ids = colData(sce)$sampletype, use.assay.type = "X")
})
# Combine
all_counts <- do.call(cbind, lapply(pb_list, assay))
all_colData <- do.call(rbind, lapply(pb_list, colData))
pb <- SingleCellExperiment(assays=list(counts=all_counts), colData=all_colData)
```


# Saving objects

Pseudobulk
```{r}
# save(pb, file = "/mnt/europa/valerio/data/Rdata/pseudobulk_nuclei.RData")
# save(sce_merged, file = "/mnt/europa/valerio/data/Rdata/sce_merged_nuclei.RData")
# save(sce_list, file = "/mnt/europa/valerio/data/Rdata/sce_list_nuclei.RData")
```

