---
title: "DGE pipeline"
output: html_document
params:
  saves_path_stem: "mia_segment"
  #labels_key: "labels_singler"
  #spe_key: "_spe_nocell_expans"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(dplyr)
library(tidyr)
library(DropletUtils)
library(SingleCellExperiment)
library(SpatialExperiment)
library(scuttle)
library(edgeR)
library(openxlsx)
library(data.table)
source("/mnt/europa/valerio/repositories/cachetic_visiumHD/r_scripts/utils/utils_fun.R")
```

# Differential gene expression analysis

Define the paths
```{r}
pl_path = paste0("/mnt/europa/valerio/repositories/cachetic_visiumHD/figures/output_R/DGE_singleR/", params$saves_path_stem,"/")
Rd_path = "/mnt/europa/valerio/data/Rdata/"
ref_path = "/mnt/europa/valerio/data/multiome_sce/multiome_sce.rds"
res_path = paste0("/mnt/europa/valerio/repositories/cachetic_visiumHD/results/DGE_singleR/", params$saves_path_stem, "/")
```

Loading pseudobulk with cell_types and contrast list
```{r}
pb_ct <- readRDS(paste0(Rd_path, "pseudobulk_celltype_SingleR"))
contrast_list <- readRDS(paste0(Rd_path, "contrast_list"))
```

```{r eval=FALSE, include=FALSE}
pb_ct
colData(pb_ct)
pb_ct$cell_type <- pb_ct$labels_singler
```


# DEGs FOR ALL SAMPLE TYPES

```{r}
deg_fun_emma <- \(pb, labels_col = "labels_singler" , contrast_list){
  lrt_list <- list()
  summary_list <- list()
  pseudobulk_exp <- pb
  pseudobulk_exp$cell_type <- colData(pseudobulk_exp)[[labels_col]]

  for (contrast_name in names(contrast_list)) {
    lrt_list[[contrast_name]] <- list()
    
    for (current_ct in unique(pseudobulk_exp$cell_type)) {
      ct_pb <- pseudobulk_exp[, pseudobulk_exp$cell_type == current_ct]
      y <- edgeR::DGEList(assay(ct_pb), samples = as.data.frame(colData(ct_pb)))
      keep <- edgeR::filterByExpr(y, group = y$samples$tissue_type)
      y <- y[keep,, keep.lib.sizes=FALSE]
      y <- edgeR::calcNormFactors(y, method="TMM")  # default
      design <- model.matrix(~ 0 + tissue_type, y$samples)
      y <- estimateDisp(y, design)
      fit <- glmFit(y, design)
      
      contrast <- makeContrasts(contrasts = contrast_list[[contrast_name]], levels = design)
      
      lrt <- glmLRT(fit, contrast = contrast)
      res <- topTags(lrt, n = Inf)$table
      res$significant <- res$FDR < 0.05
      
      sig <- res[res$significant, ]
      sig_up <- sig[sig$logFC >= 0, ]
      sig_down <- sig[sig$logFC < 0, ]
      
      lrt_list[[contrast_name]][[current_ct]] <- list(
        full = res,
        sig = sig,
        up = sig_up,
        down = sig_down
      )
      
      gene_summary <- data.table(
        contrast = contrast_name,
        cell_type = current_ct,
        Geni_considerati = nrow(res),
        Geni_significativi = nrow(sig),
        Upregolati = nrow(sig_up),
        Downregolati = nrow(sig_down)
      )
      
      summary_list[[length(summary_list) + 1]] <- gene_summary
    }
  }
  final_summary <- do.call(rbind, summary_list)
  
  return(list(summary = final_summary, lrt = lrt_list))
}
```


```{r}
# Tabella riassuntiva

degged <- deg_fun_emma(pb_ct, "labels_singler", contrast_list)

save(degged, file = paste0(Rd_path, "deg_results_singleR_list_mia_segmentazione"))

summary_df <- degged$summary
print(summary_df)
```

Summary summmary table
```{r}
final_summary_table <- summary_df |> 
  group_by(contrast) |> 
  summarise(
    N_up_DEG = sum(Upregolati, na.rm = TRUE),
    N_down_DEG = sum(Downregolati, na.rm = TRUE)
  ) |> 
  mutate(
    N_DEG = N_up_DEG + N_down_DEG
  ) |> 
  dplyr::rename(
    Confronti = contrast
  )

# Print the final table
print(final_summary_table)
```


## Summary results

```{r}

wb <- createWorkbook()

# Aggiungo uno sheet per ogni contrasto con la tabella completa
for (contrast_name in unique(summary_df$contrast)) {
  contrast_table <- subset(summary_df, contrast == contrast_name)
  
  addWorksheet(wb, contrast_name)
  writeData(wb, contrast_name, contrast_table)
}

saveWorkbook(wb = wb, file = paste0(res_path, "summary_contrasts.xlsx"), overwrite = TRUE)
```

## All results

For each contrast pvalues of every gene

```{r}
lrt_list <- degged$lrt

for (contrast_name in names(lrt_list)) {
  wb <- createWorkbook()
  
  for (cell_type in names(lrt_list[[contrast_name]])) {
    res     <- lrt_list[[contrast_name]][[cell_type]]$full
    sig     <- lrt_list[[contrast_name]][[cell_type]]$sig
    sig_up  <- lrt_list[[contrast_name]][[cell_type]]$up
    sig_down<- lrt_list[[contrast_name]][[cell_type]]$down
    
    # nome foglio pulito
    clean_name <- gsub("[^[:alnum:]_]", "_", cell_type)
    
    # foglio per cell_type con sottosezioni
    addWorksheet(wb, paste0(clean_name, "_all"))
    writeData(wb, paste0(clean_name, "_all"), res, rowNames = TRUE)
    
    addWorksheet(wb, paste0(clean_name, "_sig"))
    writeData(wb, paste0(clean_name, "_sig"), sig, rowNames = TRUE)
    
    addWorksheet(wb, paste0(clean_name, "_up"))
    writeData(wb, paste0(clean_name, "_up"), sig_up, rowNames = TRUE)
    
    addWorksheet(wb, paste0(clean_name, "_down"))
    writeData(wb, paste0(clean_name, "_down"), sig_down, rowNames = TRUE)
  }
  
  # salva il file excel del contrasto
  clean_contrast <- gsub("[^[:alnum:]_]", "_", contrast_name)
  saveWorkbook(wb, file = paste0(res_path,"dge_results/", clean_contrast, ".xlsx"), overwrite = TRUE)
}

```

#-------------------------------------------------------------------------------

# Comparison between Myo Trim63 from c26 samples and Myo IIb from sham samples

```{r}
pb_ct$group <- paste0(pb_ct$labels_singler,"_",pb_ct$tissue_type)
pb_sub <- pb_ct[, pb_ct$group %in% c("Myonuclei_Trim63_c26", "Myonuclei_IIb_sham")]
pb_sub$group <- factor(pb_sub$group, levels = c("Myonuclei_IIb_sham", "Myonuclei_Trim63_c26"))

y <- DGEList(assay(pb_sub, 'X'), samples = as.data.frame(colData(pb_sub)))
keep <- filterByExpr(y, group = y$samples$group)
y <- y[keep,, keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
design <- model.matrix(~ group, data=y$samples)
y <- estimateDisp(y, design)
fit <- glmFit(y, design)
lrt <- glmLRT(fit, coef = 2)
res <- topTags(lrt, n = Inf)$table
res$significant <- res$FDR < 0.05
sig <- res[res$significant, ]
sig_up <- sig[sig$logFC >= 0, ]
sig_down <- sig[sig$logFC < 0, ]
  
wb <- createWorkbook()
addWorksheet(wb, "Tutti_geni")
writeData(wb, "Tutti_geni", res, rowNames = TRUE)
addWorksheet(wb, "Significativi")
writeData(wb, "Significativi", sig, rowNames = TRUE)
addWorksheet(wb, "Upregolati")
writeData(wb, "Upregolati", sig_up, rowNames = TRUE)
addWorksheet(wb, "Downregolati")
writeData(wb, "Downregolati", sig_down, rowNames = TRUE)
saveWorkbook(wb, file = paste0(res_path,"dge_results/trim63_c26_vs_IIb_sham.xlsx"), overwrite = TRUE)
  
gene_summary <- data.frame(
    Geni_considerati = nrow(res),
    Geni_significativi = nrow(sig),
    Upregolati = nrow(sig_up),
    Downregolati = nrow(sig_down)
)
  
gene_summary
  
```

# ------------------------------------------------------------------------------

# Comparison between Myo Trim63 from c26 samples and Myonuclei overall from sham samples

Remodelling the pseudobulk experiment with only c26 and sham and aggregating all myonucleis
```{r}
pb_ct$group <- paste0(pb_ct$labels_singler,"_",pb_ct$tissue_type)
mio_ov_sham <- unique(grep(paste0("^", "Myonuclei", ".*", "sham"), pb_ct$group, value = TRUE))
pb_sub <- pb_ct[, pb_ct$group %in% c("Myonuclei_Trim63_c26", mio_ov_sham)]

pb_sub$group[pb_sub$group %in% mio_ov_sham] <- "Myonuclei_sham"
pb_sub$group <- as.factor(pb_sub$group)
pb_comb <- muscat::aggregateData(pb_sub, assay = "X", fun = "sum", by = "sample_id")
colData(pb_comb)$group <- pb_sub$group[match(colnames(pb_comb), pb_sub$sample_id)]
names(assays(pb_comb)) <- "counts"
```

Da notare che il blocco6 c26 ha libsize piu piccola di blocco7 c26 (quasi 4 a 1)

```{r}
y <- DGEList(counts(pb_comb), samples = as.data.frame(colData(pb_comb)))
keep <- filterByExpr(y, group = y$samples$group)
y <- y[keep,, keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
design <- model.matrix(~ group, data=y$samples)
y <- estimateDisp(y, design)
fit <- glmFit(y, design)
lrt <- glmLRT(fit, coef = 2)
top <- topTags(lrt, n = Inf)$table
top$SYMBOL <- rownames(top)

top$significant <- top$FDR < 0.05
sig <- top[top$significant, ]
sig_up <- sig[sig$logFC >= 0, ]
sig_down <- sig[sig$logFC < 0, ]
  
wb <- createWorkbook()
addWorksheet(wb, "Tutti_geni")
writeData(wb, "Tutti_geni", sig, rowNames = TRUE)
addWorksheet(wb, "Significativi")
writeData(wb, "Significativi", sig, rowNames = TRUE)
addWorksheet(wb, "Upregolati")
writeData(wb, "Upregolati", sig_up, rowNames = TRUE)
addWorksheet(wb, "Downregolati")
writeData(wb, "Downregolati", sig_down, rowNames = TRUE)
saveWorkbook(wb, file = paste0(res_path,"dge_results/trim63_c26_vs_Myo_overall_sham.xlsx"), overwrite = TRUE)

gene_summary <- data.frame(
    Geni_considerati = nrow(top),
    Geni_significativi = nrow(sig),
    Upregolati = nrow(sig_up),
    Downregolati = nrow(sig_down)
)
  
gene_summary
```

